### 2022-05-24 rendu/résultats_brouillon/r_brouillon.tex
kalman OK + ajout tex

\documentclass[titlepage]{article}
\usepackage[utf8]{inputenc}
\usepackage[french]{babel}
\usepackage[margin=1in]{geometry}
\usepackage[T1]{fontenc}
\usepackage[colorlinks=true,linkcolor=teal]{hyperref}
\usepackage{amssymb,amsmath,minted,float,graphicx,textcomp,systeme,listings,physics,lmodern,mathtools,ifthen,esvect,hyperref,fancyhdr,lastpage,marginnote,chngcntr,fancyvrb,subcaption}
\DeclareMathOperator{\e}{e}
\def\mathbi#1{\textbf{\em #1}}

\setlength\parindent{0pt}

%%%%%%%%%%%% PARAMETRES
\newcommand{\UE}{Stage}
\newcommand{\type}{Résultats}%TD-TP-COURS
\newcommand{\nb}{0}%nb 
\newcommand{\sbt}{Reconstruction dynamique d'images radioastronomiques}%soustitre
\author{Gabriel ROBERT-DAUTUN}
\date{2022}

% mise en page (header, compteur, fancy)
\pagestyle{fancy}
\lhead{\UE\, - \type\ifthenelse{\nb > 0}{\nb}{}}
\rfoot{\thepage /\pageref{LastPage}}
\lfoot{}
\cfoot{}
\renewcommand{\footrulewidth}{0.4pt}

%reset compteur section par partie
\counterwithin*{section}{part}

% permet notes + simples en marge à gauche (ig compteur de question)
\reversemarginpar
\newcommand{\mgn}[1]{\marginnote{#1}}

%sinon warning mdr
\setlength{\headheight}{13.07225pt}

%compteur questions
\newcounter{question}
\setcounter{question}{1}
\newcounter{subq}
\setcounter{subq}{1}

\newcommand{\rsubq}{\setcounter{subq}{1}}
\newcommand{\question}{\mgn{\thequestion .}\stepcounter{question}\rsubq}
\newcommand{\rstq}{\setcounter{question}{1}\rsubq}
\newcommand{\skipq}[1]{\addtocounter{question}{#1}\rsubq}
\newcommand{\subq}{\noindent\alph{subq})\stepcounter{subq}} %a modifier pour mettre dans le margin

\title{%
	\UE\, - \type\ifthenelse{\nb > 0}{\nb}{} \\
	\large \sbt}

\begin{document}

	\begin{figure}
		\centering
		\includegraphics[width=0.7\linewidth]{logo.PNG}
		\label{fig:logo}
	\end{figure}
	\maketitle

	\newpage
	\tableofcontents

	\newpage
	\section{Robustesse et performances}
	\subsection{Erreur sur l'image de départ}
	On introduit une erreur gaussienne sur l'image $X_0$. On observe que le fonctionnement du Kalman est bien celui attendu dans ce cas là : l'erreur dans l'image se réduit petit a petit.

	\begin{figure}[H]
		\centering
		\includegraphics[width=0.9\linewidth]{Perfs/Illustration.JPG}
		\caption{En haut : images réelles utilisées pour générer les données. En bas : images reconstruites. Titres : PSNR}
	\end{figure}

	On utilise une transformation de rotation de l'image de 90°. On remarque grâce au PSNR que les images reconstruites s'approchent des images réelles.

\end{document}

\begin{figure}[H]
	\centering
	\includegraphics[width=0.7\linewidth]{}
	\caption{}
\end{figure}

\begin{figure}[H]
	\centering
	\begin{subfigure}{.5\textwidth}
		\centering
		\includegraphics[width=.4\linewidth]{image1}
		\caption{}
	\end{subfigure}%
	\begin{subfigure}{.5\textwidth}
		\centering
		\includegraphics[width=.4\linewidth]{image2}
		\caption{}
	\end{subfigure}
	\caption{}
\end{figure}

### 2022-05-24 Kalman/KalmainV4.m

@@ -13,11 +13,14 @@ Nreal = 1000;
 [nY,nR,nQ,tX] = dataGen_im(H,J,N,Pix,true_image,A,Nreal,SNR_visibilites,RATIO);
 
 % génération image initiale
-P_0 = zeros(Pix,Pix);
 % ajout bruit image
 poids = 0.2;
 X_0 = true_image + randn(size(true_image))*poids;
+X_0 = X_0/max(X_0(:));
 tX(:,1) = X_0;
+
+P_0 = X_0*X_0';
+% P_0 = eye(Pix);
 
 %%%%%%%%%%%% Kalman
 X=-1;

### 2022-05-24 Kalman/Kalman_CPU_V2.m

@@ -15,7 +15,7 @@ P = zeros(Pix,Pix,n);
 %%% INITIALISATION
 X(:,1) = X_0;
 P(:,:,1) = P_0;
-
+% figure();
 for k=2:n
     fprintf('itération : %u \n',k);
     %%% PREDICTION
@@ -25,7 +25,9 @@ for k=2:n
     %%% KALMAN GAIN
     K = Pp*(H.')*pinv(H*Pp*(H.') + nR(:,k));
     %%% ESTIMATION
-    X(:,k) = Xp + K*(nY(:,k)-H*Xp);
+    X(:,k) = Xp + K*(nY(:,k)-yp);
     P(:,:,k) = Pp - K*H*Pp;
+%     subplot(n-1,1,k-1);
+%     imshow(abs(K));colorbar;axis image;colormap("parula");title(mean(K, 'all'))
 end
 end

### 2022-05-24 Kalman/dataGen_im.m

@@ -18,8 +18,8 @@ for k=1:N
         nX(:,k) = A*nX(:,k-1);
         b = sqrt(mean(nX(:,k).^2,'all')/RATIO)*randn(size(nX(:,k))); % erreur evolution
     end
-    xb = abs(nX(:,k)+b);
-    b = nX(:,k)-xb;
+%     xb = abs(nX(:,k)+b);
+%     b = nX(:,k)-xb;
     b = zeros(size(b));
     nQ(:,:,k) = (b*b')/Pix;
 

### 2022-05-24 Kalman/dispKalman_t.m

@@ -4,7 +4,7 @@ function  dispKalman_t(X,tX,N,Mx,My)
 figure();
 for k=1:N
     subplot(2,N,k);imshow(y_to_im(abs(tX(:,k)),Mx,My));colorbar;axis image;colormap("parula")
-    subplot(2,N,k+N);imshow(y_to_im(abs(X(:,k)),Mx,My));colorbar;axis image;colormap("parula");title(psnr(X(:,k),tX(:,k)))
+    subplot(2,N,k+N);imshow(y_to_im(abs(X(:,k)),Mx,My));colorbar;axis image;colormap("parula");title(psnr(abs(X(:,k)),tX(:,k)))
 end
 end
 

### 2022-05-24 rendu/résultats_brouillon/r_brouillon.tex
err modèle

@@
 
-	On utilise une transformation de rotation de l'image de 90°. On remarque grâce au PSNR que les images reconstruites s'approchent des images réelles.
+	On utilise une transformation de rotation de l'image de 90°.
+On remarque grâce au PSNR que les images reconstruites s'approchent des images réelles, mais pas parfaitement car le filtre prend en compte l'historique.
+Une erreur dans une image va donc se diffuser aux images suivantes (précédentes si backwards).
+On observe la convergence de reconstruction :
+
+	\begin{figure}[H]
+		\centering
+		\includegraphics[width=0.9\linewidth]{Perfs/conv02.png}
+		\caption{Evolution psnr en fonction du nombre d'itérations}
+	\end{figure}
+
+	On converge en 4 itérations au plus a chaque essai : les erreurs gaussienne d'amplitude 0.2 se diffusent sur 4 itérations environ.
+	Vérifié pour des amplitude de 0.1 a 1. Pour des amplitudes plus grandes (5 donc SNR=0.1) parfois 5 itérations sont nécessaires
 

### 2022-05-24 main_2.m

@@ -11,7 +11,7 @@ c= 3e8;
 f = 100e6;
 lambda = c/f;
 dist = lambda/2.1; % J_b antennes espacées de dist (normalisé)
-N = 5;
+N = 25;
 SNR = 10;
 iMEM = 1; % nombre itérations MEM 
 RATIO = 10e14; % ratio erreur Kalman
@@ -39,4 +39,4 @@ A_ev(7,1) = 1; A_ev(4,2)=1; A_ev(1,3)=1; A_ev(8,4)=1;A_ev(5,5)=1;A_ev(2,6)=1;A_e
 
 %%% Kalman direct en fait mdr
 [X,tX] = KalmainV4(vadapted,J,z,I,N,f,c,lambda,Mx,My,iMEM,SNR,A_ev,RATIO,mode);
-dispKalman_t(X,tX,N,Mx,My);
+% dispKalman_t(X,tX,N,Mx,My);

### 2022-05-24 Kalman/KalmainV4.m
!! retirer z_err + based

@@ -14,7 +14,7 @@ Nreal = 1000;
 
 % génération image initiale
 % ajout bruit image
-poids = 0.01;
+poids = 1e-15;
 X_0 = true_image ;
 tX(:,1) = X_0;
 
@@ -23,12 +23,12 @@ P_0 = X_0*X_0';
 % P_0 = eye(Pix);
 
 %ajout erreur sur A (!!!APRES!!! dataGen => sinon c'est pas une erreur)
-A_err = abs(A + poids*randn(size(A)));
+H_err = abs(H + poids*randn(size(H)));
 
 %%%%%%%%%%%% Kalman
 X=-1;
 if mode=='CPU'
-    X = Kalman_CPU_V2(A_err,H,X_0,P_0,nY,nR,nQ,Pix,N);
+    X = Kalman_CPU_V2(A,H_err,X_0,P_0,nY,nR,nQ,Pix,N);
 end
 if mode=='GPU'
     X = Kalman_GPU_V4(A,H,X_0,P_0,nY,nR,nQ,Pix,N);

### 2022-05-24 Kalman/dispKalman_t.m

@@ -4,7 +4,7 @@ function  dispKalman_t(X,tX,N,Mx,My)
 figure();
 for k=1:N
     subplot(2,N,k);imshow(y_to_im(abs(tX(:,k)),Mx,My));colorbar;axis image;colormap("parula")
-    subplot(2,N,k+N);imshow(y_to_im(abs(X(:,k)),Mx,My));colorbar;axis image;colormap("parula");title(psnr(abs(X(:,k)),tX(:,k)))
+    subplot(2,N,k+N);imshow(y_to_im(abs(X(:,k)/max(abs(X(:,k)))),Mx,My));colorbar;axis image;colormap("parula");title(psnr(abs(X(:,k)/max(abs(X(:,k)))),tX(:,k)))
 end
 end
 

### 2022-05-24 main_2.m

@@ -40,3 +40,7 @@ A_ev(7,1) = 1; A_ev(4,2)=1; A_ev(1,3)=1; A_ev(8,4)=1;A_ev(5,5)=1;A_ev(2,6)=1;A_e
 %%% Kalman direct en fait mdr
 [X,tX] = KalmainV4(vadapted,J,z,I,N,f,c,lambda,Mx,My,iMEM,SNR,A_ev,RATIO,mode);
 % dispKalman_t(X,tX,N,Mx,My);
+
+z_err = z+ randn(size(z))*0.01;
+H = matF(J,Pix,z,lambda,I);
+H_err = matF(J,Pix,z_err,lambda,I);

### 2022-05-30 rendu/résultats_brouillon/r_brouillon.tex
!! retirer z_err + based
cant write ?

@@
 	\newpage
-	\section{Robustesse et performances}
+	\section{Influence individuelle de l'erreur sur les matrices}
 	\subsection{Erreur sur l'image de départ}
@@
 	On converge en 4 itérations au plus a chaque essai : les erreurs gaussienne d'amplitude 0.2 se diffusent sur 4 itérations environ.
-	Vérifié pour des amplitude de 0.1 a 1. Pour des amplitudes plus grandes (5 donc SNR=0.1) parfois 5 itérations sont nécessaires
+	Vérifié pour des amplitude de 0.1 a 1. Pour des amplitudes plus grandes (5 donc SNR=0.1) parfois 5 itérations sont nécessaires.
+
+	\subsection{Erreur sur les matrices de covariance}
+
+	Les matrices $\{P_k\}$ et $\{Q_k\}$ servent à calculer le gain de Kalman et donc sélectionner quelle proportion mesure/prédiction choisir pour estimer l'image.
+Si on introduit une erreur \emph{uniquement} sur une de ces matrices, l'image prédite et l'image mesurée seront les mêmes (car on y a pas introduit d'erreur) et donc une erreur sur la proportion n'aura aucun effet.\\
+	Une erreur sur ces matrices n'introduit donc pas d'erreur sur la reconstruction de manière directe. Elle peut néanmoins diffuser une erreur déjà présente.
+
+	\subsection{Erreur sur le modèle d'évolution}
 

### 2022-06-01 Kalman/KalmainV4.m
jsais meme pas mdr

@@ -28,7 +28,7 @@ H_err = abs(H + poids*randn(size(H)));
 %%%%%%%%%%%% Kalman
 X=-1;
 if mode=='CPU'
-    X = Kalman_CPU_V2(A,H_err,X_0,P_0,nY,nR,nQ,Pix,N);
+    X = Kalman_CPU_V2(A,H,X_0,P_0,nY,nR,nQ,Pix,N);
 end
 if mode=='GPU'
     X = Kalman_GPU_V4(A,H,X_0,P_0,nY,nR,nQ,Pix,N);
@@ -39,6 +39,12 @@ for k=1:N
     Psr(k) = psnr(abs(X(:,k)),tX(:,k));
 end
 figure();plot(Psr);title('PSNR = f(it)')
+
+%{
+z_err = z+ randn(size(z))*0.01;
+H = matF(J,Pix,z,lambda,I);
+H_err = matF(J,Pix,z_err,lambda,I);
+%}
 
 end
 

### 2022-06-01 main_2.m

@@ -1,5 +1,4 @@
-addpath(genpath(pwd))
-
+addpath(genpath(pwd)) % ajoute tous les sous dossiers
 close all
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@ -8,10 +7,10 @@ Mx = 3; %My auto (ratio)
 J_b = 5; % On suppose le réseau d'antennes carré de côté
 % max 11 GPU atm
 c= 3e8;
-f = 100e6;
+f = 1e9;
 lambda = c/f;
-dist = lambda/2.1; % J_b antennes espacées de dist (normalisé)
-N = 25;
+dist = 100*lambda/2.1; % J_b antennes espacées de dist (normalisé)
+N = 9;
 SNR = 10;
 iMEM = 1; % nombre itérations MEM 
 RATIO = 10e14; % ratio erreur Kalman
@@ -39,8 +38,4 @@ A_ev(7,1) = 1; A_ev(4,2)=1; A_ev(1,3)=1; A_ev(8,4)=1;A_ev(5,5)=1;A_ev(2,6)=1;A_e
 
 %%% Kalman direct en fait mdr
 [X,tX] = KalmainV4(vadapted,J,z,I,N,f,c,lambda,Mx,My,iMEM,SNR,A_ev,RATIO,mode);
-% dispKalman_t(X,tX,N,Mx,My);
-
-z_err = z+ randn(size(z))*0.01;
-H = matF(J,Pix,z,lambda,I);
-H_err = matF(J,Pix,z_err,lambda,I);
+dispKalman_t(X,tX,N,Mx,My);

### 2022-06-01 sim/matF.m

@@ -12,7 +12,7 @@ end
 dz = reshape(dz,J^2,2);
 
 for q=1:Q
-      F(: ,q) = exp(-1i*2*pi/lambda*(dz(:,1)*I(q,1)+dz(:,2)*I(q,2)));
-end
+    F(: ,q) = exp(-1i*2*pi/lambda*(dz(:,1)*I(q,1)+dz(:,2)*I(q,2)));
+end % .* aussi non ?
 end
 

### 2022-06-02 rendu/résultats_brouillon/r_brouillon.tex
ltx ok
ltx + err H

@@
 
+	On suppose une erreur sur le modèle d'évolution \emph{dans le filtre de Kalman}, afin de simuler un modèle théorique imprécis. On construit les données simulées avec un modèle sans erreur.\\
+	\begin{minted}{matlab}
+poids = 0.2;
+A = A + poids*randn(size(A));
+	\end{minted}
+
+	On introduit une erreur gaussienne
+
+	\subsection{Erreur sur le masque UV}
+
+	On cherche à caractériser l'erreur sur la matrice du plan UV $H$.
+Une manière naïve de faire le calcul serait d'ajouter une erreur gaussienne comme sur la matrice $A$, mais en faisant cela on obtient un modèle très instable et peu réaliste.
+Effectivement, les termes de cette matrices étant :
+	$$
+		H_{j,q} = \exp(-j\frac{2\pi}{\lambda}\Delta z_j\dotproduct\mathbf{I}_q)
+	$$
+	L'erreur sur cette matrice pouvant provenir de différentes sources : mauvaise mesure des distances entre les antennes, mauvais alignement des antennes donc erreur sur les directions, le modèle d'erreur est :
+	$$
+		H_{j,q}^{err} =  \exp(-j\frac{2\pi}{\lambda}(\Delta z_j\dotproduct\mathbf{I}_q + \epsilon)) =  \exp(-j\frac{2\pi}{\lambda}\Delta z_j\dotproduct\mathbf{I}_q)\exp(-j\frac{2\pi}{\lambda}\epsilon)
+	$$
+
+	Soit une rotation du complexe associé a chaque direction. \\
+
+	On implémente donc l'erreur de la manière suivante pour quantifier l'erreur acceptable sur les positions des antennes :
+
+	\begin{minted}{matlab}
+poids = 1e-2;
+z_err = z+ randn(size(z))*poids;
+H_err = matF(J,Pix,z_err,lambda,I);
+	\end{minted}
+
+	On peut ensuite mesurer l'évolution du PSNR dans le temps, pour différents écarts possibles :
+
+%	\begin{figure}[H]
+%		\centering
+%		\includegraphics[width=0.7\linewidth]{Perfs/H5e-3}
+%		\caption{Evolution du PSNR pour $<\epsilon^2>=5$mm}
+%		\label{fig:h5e-3}
+%	\end{figure}
+%	
+%	\begin{figure}[H]
+%		\centering
+%		\includegraphics[width=0.7\linewidth]{Perfs/He-3_1}
+%		\caption{Evolution du PSNR pour $<\epsilon^2>=1$mm}
+%		\label{fig:he-3}
+%	\end{figure}
+%	
+%	\begin{figure}[H]
+%		\centering
+%		\includegraphics[width=0.7\linewidth]{Perfs/He-2}
+%		\caption{Evolution du PSNR pour $<\epsilon^2>=1$cm}
+%		\label{fig:he-2}
+%	\end{figure}
+
+	\begin{figure}[H]
+		\centering
+		\begin{subfigure}{.5\textwidth}
+			\centering
+			\includegraphics[width=0.9\linewidth]{Perfs/He-3_1}
+			\caption{Evolution du PSNR pour $<\epsilon^2>=1$mm}
+			\label{fig:he-3}
+		\end{subfigure}%
+		\begin{subfigure}{.5\textwidth}
+			\centering
+			\includegraphics[width=0.9\linewidth]{Perfs/H5e-3}
+			\caption{Evolution du PSNR pour $<\epsilon^2>=5$mm}
+			\label{fig:h5e-3}
+		\end{subfigure}
+		\begin{subfigure}{.5\textwidth}
+			\centering
+			\includegraphics[width=0.9\linewidth]{Perfs/He-2}
+			\caption{Evolution du PSNR pour $<\epsilon^2>=1$cm}
+			\label{fig:he-2}
+		\end{subfigure}
+		\caption{Mesure du psnr pour différentes erreurs sur la position}
+	\end{figure}
+
+	Les mesures obtenues ont \emph{généralement} cette forme : on obtient parfois des évolutions sans le pic au début, mais la mesure va toujours converger vers une valeur identique tant que $<\epsilon^2>$ reste constant, en oscillant autour de cette valeur.
+On peut donc tracer l'évolution de l'erreur de reconstruction en fonction de $<\epsilon^2>$ :
+
 \end{document}
@@
 		\centering
-		\includegraphics[width=.4\linewidth]{image1}
+		\includegraphics[width=.9\linewidth]{image1}
 		\caption{}
@@
 		\centering
-		\includegraphics[width=.4\linewidth]{image2}
+		\includegraphics[width=.9\linewidth]{image2}
 		\caption{}

### 2022-06-02 main_2.m

@@ -10,7 +10,7 @@ c= 3e8;
 f = 1e9;
 lambda = c/f;
 dist = 100*lambda/2.1; % J_b antennes espacées de dist (normalisé)
-N = 9;
+N = 15;
 SNR = 10;
 iMEM = 1; % nombre itérations MEM 
 RATIO = 10e14; % ratio erreur Kalman
@@ -38,4 +38,4 @@ A_ev(7,1) = 1; A_ev(4,2)=1; A_ev(1,3)=1; A_ev(8,4)=1;A_ev(5,5)=1;A_ev(2,6)=1;A_e
 
 %%% Kalman direct en fait mdr
 [X,tX] = KalmainV4(vadapted,J,z,I,N,f,c,lambda,Mx,My,iMEM,SNR,A_ev,RATIO,mode);
-dispKalman_t(X,tX,N,Mx,My);
+% dispKalman_t(X,tX,N,Mx,My);

### 2022-06-02 Kalman/KalmainV4.m
ça fait des mesures

@@ -14,7 +14,7 @@ Nreal = 1000;
 
 % génération image initiale
 % ajout bruit image
-poids = 1e-2;
+poids = 5;
 X_0 = true_image ;
 tX(:,1) = X_0;
 
@@ -41,7 +41,8 @@ for k=1:N
     %     Psr(k) = psnr(abs(X(:,k)),tX(:,k));
     Psr(k) = psnr(abs(X(:,k)/max(abs(X(:,k)))),tX(:,k));
 end
-figure();plot(Psr);title('PSNR = f(it)')
+% figure();plot(Psr);title('PSNR = f(it)')
+mean(Psr(6:end))
 
 %{
 z_err = z+ randn(size(z))*0.01;

### 2022-06-07 rendu/résultats_brouillon/r_brouillon.tex
font to fourier <3

@@
 \documentclass[titlepage]{article}
+\usepackage[upright]{fourier}
 \usepackage[utf8]{inputenc}
@@
 \usepackage[colorlinks=true,linkcolor=teal]{hyperref}
-\usepackage{amssymb,amsmath,minted,float,graphicx,textcomp,systeme,listings,physics,lmodern,mathtools,ifthen,esvect,hyperref,fancyhdr,lastpage,marginnote,chngcntr,fancyvrb,subcaption}
+\usepackage{amssymb,amsmath,minted,float,graphicx,textcomp,systeme,listings,physics,mathtools,ifthen,esvect,hyperref,fancyhdr,lastpage,marginnote,chngcntr,fancyvrb,subcaption}
+\usepackage{stmaryrd}
 \DeclareMathOperator{\e}{e}
@@
 	$$
-		H_{j,q}^{err} =  \exp(-j\frac{2\pi}{\lambda}(\Delta z_j\dotproduct\mathbf{I}_q + \epsilon)) =  \exp(-j\frac{2\pi}{\lambda}\Delta z_j\dotproduct\mathbf{I}_q)\exp(-j\frac{2\pi}{\lambda}\epsilon)
+		H_{j,q}^{err} =  \exp(-i\frac{2\pi}{\lambda}(\Delta z_j\dotproduct\mathbf{I}_q + \varepsilon)) =  \exp(-i\frac{2\pi}{\lambda}\Delta z_j\dotproduct\mathbf{I}_q)\exp(-i\frac{2\pi}{\lambda}\varepsilon\dotproduct\mathbf{I}_q)
 	$$
@@
 %		\includegraphics[width=0.7\linewidth]{Perfs/H5e-3}
-%		\caption{Evolution du PSNR pour $<\epsilon^2>=5$mm}
+%		\caption{Evolution du PSNR pour $<\varepsilon^2>=5$mm}
 %		\label{fig:h5e-3}
@@
 %		\includegraphics[width=0.7\linewidth]{Perfs/He-3_1}
-%		\caption{Evolution du PSNR pour $<\epsilon^2>=1$mm}
+%		\caption{Evolution du PSNR pour $<\varepsilon^2>=1$mm}
 %		\label{fig:he-3}
@@
 %		\includegraphics[width=0.7\linewidth]{Perfs/He-2}
-%		\caption{Evolution du PSNR pour $<\epsilon^2>=1$cm}
+%		\caption{Evolution du PSNR pour $<\varepsilon^2>=1$cm}
 %		\label{fig:he-2}
@@
 			\includegraphics[width=0.9\linewidth]{Perfs/He-3_1}
-			\caption{Evolution du PSNR pour $<\epsilon^2>=1$mm}
+			\caption{Evolution du PSNR pour $<\varepsilon^2>=1$mm}
 			\label{fig:he-3}
@@
 			\includegraphics[width=0.9\linewidth]{Perfs/H5e-3}
-			\caption{Evolution du PSNR pour $<\epsilon^2>=5$mm}
+			\caption{Evolution du PSNR pour $<\varepsilon^2>=5$mm}
 			\label{fig:h5e-3}
@@
 			\includegraphics[width=0.9\linewidth]{Perfs/He-2}
-			\caption{Evolution du PSNR pour $<\epsilon^2>=1$cm}
+			\caption{Evolution du PSNR pour $<\varepsilon^2>=1$cm}
 			\label{fig:he-2}
@@
 
-	Les mesures obtenues ont \emph{généralement} cette forme : on obtient parfois des évolutions sans le pic au début, mais la mesure va toujours converger vers une valeur identique tant que $<\epsilon^2>$ reste constant, en oscillant autour de cette valeur.
-On peut donc tracer l'évolution de l'erreur de reconstruction en fonction de $<\epsilon^2>$ :
+	Les mesures obtenues ont \emph{généralement} cette forme : on obtient parfois des évolutions sans le pic au début, mais la mesure va toujours converger vers une valeur identique tant que $<\varepsilon^2>$ reste constant, en oscillant autour de cette valeur.
+On peut donc tracer l'évolution de l'erreur de reconstruction en fonction de $<\varepsilon^2>$ :
+
+	\begin{figure}[H]
+		\centering
+		\includegraphics[width=0.7\linewidth]{Perfs/perf_errF}
+		\caption{Qualité de reconstruction en fonction de l'erreur}
+	\end{figure}
+
+	\subsubsection{Calculs théoriques}
+
+	On introduit la notation $\odot$ associée au produit de Hadamard, qui effectue la multiplication élément à élément de deux matrices de même taille :
+	$$
+		\forall (A,B)\in\left(\mathbb{C}^{m\times n}\right)^2, \quad A\odot B\in\mathbb{C}^{m\times n} \quad\text{et}\quad \forall (i,j)\in\llbracket1;m\rrbracket\times\llbracket1;n\rrbracket,\, (A\odot B)_{i,j} = A_{i,j}\times B_{i,j}
+	$$
+
+	On considère la matrice $H$ construite avec une erreur de mesure sur la position des antennes, c'est à dire 
+	$$
+		\forall j \in \llbracket1;J^2\rrbracket,\, \widetilde{\Delta} z_j = \Delta z_j + \boldsymbol{\varepsilon}_j
+	$$
 

### 2022-06-07 Kalman/KalmainV4.m

@@ -14,7 +14,7 @@ Nreal = 1000;
 
 % génération image initiale
 % ajout bruit image
-poids = 5;
+poids = 0.14;
 X_0 = true_image ;
 tX(:,1) = X_0;
 
@@ -39,7 +39,7 @@ end
 Psr = zeros(N,1);
 for k=1:N
     %     Psr(k) = psnr(abs(X(:,k)),tX(:,k));
-    Psr(k) = psnr(abs(X(:,k)/max(abs(X(:,k)))),tX(:,k));
+    Psr(k) = ssim(abs(X(:,k)/max(abs(X(:,k)))),tX(:,k));
 end
 % figure();plot(Psr);title('PSNR = f(it)')
 mean(Psr(6:end))

### 2022-06-07 Kalman/dispKalman_t.m

@@ -4,7 +4,7 @@ function  dispKalman_t(X,tX,N,Mx,My)
 figure();
 for k=1:N
     subplot(2,N,k);imshow(y_to_im(abs(tX(:,k)),Mx,My));colorbar;axis image;colormap("parula")
-    subplot(2,N,k+N);imshow(y_to_im(abs(X(:,k)/max(abs(X(:,k)))),Mx,My));colorbar;axis image;colormap("parula");title(psnr(abs(X(:,k)/max(abs(X(:,k)))),tX(:,k)))
+    subplot(2,N,k+N);imshow(y_to_im(abs(X(:,k)/max(abs(X(:,k)))),Mx,My));colorbar;axis image;colormap("parula");title(ssim(abs(X(:,k)/max(abs(X(:,k)))),tX(:,k)))
 end
 end
 

### 2022-06-07 main_2.m

@@ -3,14 +3,14 @@ close all
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 % PARAMETRES
-Mx = 3; %My auto (ratio)
-J_b = 5; % On suppose le réseau d'antennes carré de côté
+Mx = 5; %My auto (ratio)
+J_b = 6; % On suppose le réseau d'antennes carré de côté
 % max 11 GPU atm
 c= 3e8;
 f = 1e9;
 lambda = c/f;
-dist = 100*lambda/2.1; % J_b antennes espacées de dist (normalisé)
-N = 15;
+dist = 100*lambda/2.1; % J_b antennes espacées de dist 
+N = 7;
 SNR = 10;
 iMEM = 1; % nombre itérations MEM 
 RATIO = 10e14; % ratio erreur Kalman
@@ -19,11 +19,12 @@ mode = 'CPU';
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
 %%% choix / gen image
-% image = imread('bh_pic.jpg');
-% adapted = adaptImg(img, Mx); 
+img = imread('disques.jpg');
+adapted = adaptImg(img, Mx); 
+vadapted = vectorize(adapted);
 
-vadapted = reshape(linspace(0,1,9),9,1);
-adapted = reshape(vadapted, 3, 3);
+% vadapted = reshape(linspace(0,1,9),9,1);
+% adapted = reshape(vadapted, 3, 3);
 
 %%% Calcul paramètres
 My = size(adapted, 2);
@@ -32,10 +33,10 @@ J = size(z, 1);
 Pix = numel(adapted);
 I = matI(Mx, My);
 
-% A_ev = eye(Pix);
-A_ev = zeros(Pix);
-A_ev(7,1) = 1; A_ev(4,2)=1; A_ev(1,3)=1; A_ev(8,4)=1;A_ev(5,5)=1;A_ev(2,6)=1;A_ev(9,7)=1;A_ev(6,8)=1;A_ev(3,9)=1;
+A_ev = eye(Pix);
+% A_ev = zeros(Pix);
[…]

### 2022-06-07 rendu/résultats_brouillon/r_brouillon.tex
ça avance le latex

@@
 
-	On considère la matrice $H$ construite avec une erreur de mesure sur la position des antennes, c'est à dire 
+	On considère la matrice $\widetilde{H}$ construite avec une erreur de mesure sur la position des antennes, c'est à dire 
 	$$
@@
 	$$
+	On suppose l'erreur $\boldsymbol{\varepsilon}$ suivant une loi normale.\\
+	Les termes de la matrice $\widetilde{H}$ deviennent :
+	$$
+		\widetilde{H}_{j,q} = \exp(-i\frac{2\pi}{\lambda}\widetilde{\Delta}z_j\dotproduct\mathbf{I}_q) = H_{j,q}\times\exp(-i\frac{2\pi}{\lambda}\boldsymbol{\varepsilon_j}\dotproduct\mathbf{I}_q)
+	$$
+
+	\newcommand{\C}{\mathcal{C}}
+	On peut donc établir une matrice de correction $\mathcal{C}$ telle que :
+	$$
+		H = \mathcal{C}\odot\widetilde{H}
+	$$
+	\newpage
+
+	On peut vérifier que l'expression de $K$ reste inchangée : on exprime $P_{k|k}$ :
+	\begin{align*}
+		P_{k|k} &= E\left[\left(\widehat{x}_{k|k} - x_k\right)\left(\widehat{x}_{k|k} - x_k\right)^H\right] \\
+		&= E\left[\left(\widehat{x}_{k|k-1} + K(y_k - H\widehat{x}_{k|k-1}) - x_k\right)\left(\widehat{x}_{k|k-1} + K(y_k - H\widehat{x}_{k|k-1}) - x_k\right)^H\right] \\
+		&= E\left[\left(\widehat{x}_{k|k-1} - K(\C\odot\widetilde{H})\widehat{x}_{k|k-1} + Ky_k - x_k\right)\left(\widehat{x}_{k|k-1} - K(\C\odot\widetilde{H})\widehat{x}_{k|k-1} + Ky_k - x_k\right)^H\right] \\
+		&= E\left[\left((\widehat{x}_{k|k-1}-x_k) -K(\C\odot\widetilde{H})(\widehat{x}_{k|k-1}-x_k) + K(y_k - (\C\odot\widetilde{H})x_k)\right)\left((\widehat{x}_{k|k-1}-x_k) -K(\C\odot\widetilde{H})(\widehat{x}_{k|k-1}-x_k) + K(y_k - (\C\odot\widetilde{H})x_k)\right)^H\right]
+	\end{align*}
+
+	On peut faire apparaître les erreurs de mesure $\varepsilon_{mes}$ et de prédiction $\varepsilon_{pred}$ :
+	\begin{align*}
+		\varepsilon_{mes} &= y_k - (\C\odot\widetilde{H})x_k \\
+		\varepsilon_{pred} &= \widehat{x}_{k|k-1} - x_k
+	\end{align*}
+
+	Ces erreurs sont décorrelées, donc les termes correspondant seront nuls :
+	$$
+		E\left[\varepsilon_{mes}\varepsilon_{pred}^H\right] = 0
+	$$
+
+	En rappelant de plus :
+	\begin{align*}
+		P_{k|k-1} &= E\left[\varepsilon_{pred}\varepsilon_{pred}^H\right] \\
+		R &=  E\left[\varepsilon_{mes}\varepsilon_{mes}^H\right]
+	\end{align*}
+
+	Erreurs aux indices k, on peut développer la formule de $P_{k|k-1}
 

### 2022-06-08 rendu/résultats_brouillon/r_brouillon.tex
html fail + update

@@
 \usepackage[T1]{fontenc}
+
 \usepackage[colorlinks=true,linkcolor=teal]{hyperref}
@@
 \newcommand{\subq}{\noindent\alph{subq})\stepcounter{subq}} %a modifier pour mettre dans le margin
+
+\newcommand{\C}{\mathcal{C}}
+\newcommand{\Ht}{\widetilde{H}}
+\newcommand{\Hc}{\C\odot\Ht}
 
@@
 	$$
-		\widetilde{H}_{j,q} = \exp(-i\frac{2\pi}{\lambda}\widetilde{\Delta}z_j\dotproduct\mathbf{I}_q) = H_{j,q}\times\exp(-i\frac{2\pi}{\lambda}\boldsymbol{\varepsilon_j}\dotproduct\mathbf{I}_q)
+		\widetilde{H}_{j,q} = \exp(-i\frac{2\pi}{\lambda}\widetilde{\Delta}z_j\dotproduct\mathbf{I}_q) = H_{j,q}\times\exp(-i\frac{2\pi}{\lambda}\boldsymbol{\varepsilon}_j\dotproduct\mathbf{I}_q)
 	$$
 
-	\newcommand{\C}{\mathcal{C}}
 	On peut donc établir une matrice de correction $\mathcal{C}$ telle que :
@@
 
-	Erreurs aux indices k, on peut développer la formule de $P_{k|k-1}
+	Erreurs aux indices k, on peut développer la formule de $P_{k|k}$
 
+	\begin{align*}
+		P_{k|k} &= P_{k|k-1} + K\left(\C\odot \widetilde{H}\right)P_{k|k-1}\left(\Hc\right)^H - P_{k|k-1}\left(K\Hc\right)^H - K\left(\Hc\right)P_{k|k-1} + KRK^H
+	\end{align*}
+
+	Cette forme est bien minimisée par 
+	\begin{equation}
+		K_k^c = P_{k|k-1}\left(\Hc\right)^H\left(\left(\Hc\right)P_{k|k-1}\left(\Hc\right)^H + R\right)^{-1}
+	\end{equation}
+	Forme similaire au gain $K$ calculé sans correction :
+	\begin{equation}
+		K_k = P_{k|k-1}\Ht^H\left(\Ht P_{k|k-1}\Ht^H + R\right)^{-1}
+	\end{equation}
+
+	Par la suite, on omettra les dépendances en $k$ en supposant $P=P_{k|k-1}$ et $K = K_k$.
+
+	On pose l'erreur sur le gain :
+	\begin{equation}
+		\varepsilon_K = K^c - K
+	\end{equation}
+
+	Ainsi que l'erreur d'estimation induite par l'erreur de masque :
+	\begin{equation}
+		\varepsilon_{\widehat{x}} = \widehat{x}_{k|k} - \widehat{x}_{k|k}^c
+	\end{equation}
+
+	On omettra de la même façon la dépendance en $k$. En développant :
+	\begin{align*}
+		\varepsilon_{\widehat{x}} &= \left[\widehat{x}_{k|k-1} + K(y_k-\Ht\widehat{x}_{k|k-1})\right] - \left[\widehat{x}_{k|k-1} + K^c\left(y_k - \left(\Hc\right)\widehat{x}_{k|k-1}\right)\right] \\
+		&= \left(K-K^c\right)y_k - \left(K\Ht - K^c(\Hc)\right)\widehat{x}_{k|k-1}
+	\end{align*}
 \end{document}

### 2022-06-09 rendu/résultats_brouillon/r_brouillon.tex
update laetx

@@
 	$$
-		\forall j \in \llbracket1;J^2\rrbracket,\, \widetilde{\Delta} z_j = \Delta z_j + \boldsymbol{\varepsilon}_j
+		\forall j \in \llbracket1;J^2\rrbracket,\:\: \widetilde{\Delta} z_j = \Delta z_j + \boldsymbol{\varepsilon}_j
 	$$
@@
 		\varepsilon_{\widehat{x}} &= \left[\widehat{x}_{k|k-1} + K(y_k-\Ht\widehat{x}_{k|k-1})\right] - \left[\widehat{x}_{k|k-1} + K^c\left(y_k - \left(\Hc\right)\widehat{x}_{k|k-1}\right)\right] \\
-		&= \left(K-K^c\right)y_k - \left(K\Ht - K^c(\Hc)\right)\widehat{x}_{k|k-1}
+		&= \left(K-K^c\right)y_k - \left(K\Ht - K^c(\Hc)\right)\widehat{x}_{k|k-1} \\
+		&= \varepsilon_K\left[y_k - \left(\Hc\right)\widehat{x}_{k|k-1}\right] + K\left(\mathbb{1}-\C\right)\odot\Ht\widehat{x}_{k|k-1}
 	\end{align*}
+
+	Où $\mathbb{1}$ désigne l'élément neutre du produit de Hadamard, la matrice dont tous les termes valent 1.\\	
+	Donc :
+	\begin{equation}
+		\varepsilon_{\widehat{x}} = \varepsilon_K\varepsilon_{mes} + K\left(\mathbb{1} - \mathcal{C}\right)\odot\Ht\widehat{x}_{k|k-1}
+	\end{equation}
+
+	Sous l'hypothèse de l'introduction d'erreur uniquement sur $H$, la matrice $\C\odot\Ht$ est la matrice corrigée et on obtient :
+	$$
+		y_k = \Hc x_k \implies \varepsilon_{mes} = 0
+	$$
+	D'où :
+	\begin{equation}
+		\varepsilon_{\widehat{x}} = K \left(\mathbb{1} - \C\right)\odot\Ht\widehat{x}_{k|k-1}
+	\end{equation}
 \end{document}

### 2022-06-09 Kalman/KalmainV4.m
- dependance lambda

@@ -14,7 +14,7 @@ Nreal = 1000;
 
 % génération image initiale
 % ajout bruit image
-poids = 0.14;
+poids = 0;
 X_0 = true_image ;
 tX(:,1) = X_0;
 

### 2022-06-09 main_2.m

@@ -9,7 +9,7 @@ J_b = 6; % On suppose le réseau d'antennes carré de côté
 c= 3e8;
 f = 1e9;
 lambda = c/f;
-dist = 100*lambda/2.1; % J_b antennes espacées de dist 
+dist = 10; % J_b antennes espacées de dist 
 N = 7;
 SNR = 10;
 iMEM = 1; % nombre itérations MEM 

### 2022-06-09 sim/matF.m

@@ -12,7 +12,7 @@ end
 dz = reshape(dz,J^2,2);
 
 for q=1:Q
-    F(: ,q) = exp(-1i*2*pi/lambda*(dz(:,1)*I(q,1)+dz(:,2)*I(q,2)));
+    F(: ,q) = exp(-1i*2*pi*(dz(:,1)*I(q,1)+dz(:,2)*I(q,2)));
 end % .* aussi non ?
 end
 

### 2022-06-13 rendu/résultats_brouillon/r_brouillon.tex
rtdyguhi

@@
 \newcommand{\Hc}{\C\odot\Ht}
+\newcommand{\B}{\mathcal{B}}
+\newcommand{\Hb}{\B\odot\Ht}
+\newcommand{\hinv}[1]{#1^{\circ-1}}
 
@@
 	\tableofcontents
+
+	\newpage
+	\section{Notations}
+	\subsection{Produit de Hadamard}
+
+	On introduit la notation $\odot$ associée au produit de Hadamard, qui effectue la multiplication élément à élément de deux matrices de même taille :
+	$$
+	\forall (A,B)\in\left(\mathbb{C}^{m\times n}\right)^2, \quad A\odot B\in\mathbb{C}^{m\times n} \quad\text{et}\quad \forall (i,j)\in\llbracket1;m\rrbracket\times\llbracket1;n\rrbracket,\, (A\odot B)_{i,j} = A_{i,j}\times B_{i,j}
+	$$
+
+	On introduit également l'inverse de Hadamard, qui inverse élément par élément une matrice
+	$$
+		\forall A \in \left(\mathbb{C}\backslash\{0\}\right)^{m\times n},\quad \hinv{A}\in\left(\mathbb{C}\backslash\{0\}\right)^{m\times n} \quad\text{et}\quad \forall (i,j)\in\llbracket1;m\rrbracket\times\llbracket1;n\rrbracket,\, \left(\hinv{A}\right)_{i,j} = \left(A_{i,j}\right)^{-1}
+	$$
 
@@
 
-	On introduit la notation $\odot$ associée au produit de Hadamard, qui effectue la multiplication élément à élément de deux matrices de même taille :
-	$$
-		\forall (A,B)\in\left(\mathbb{C}^{m\times n}\right)^2, \quad A\odot B\in\mathbb{C}^{m\times n} \quad\text{et}\quad \forall (i,j)\in\llbracket1;m\rrbracket\times\llbracket1;n\rrbracket,\, (A\odot B)_{i,j} = A_{i,j}\times B_{i,j}
-	$$
-
 	On considère la matrice $\widetilde{H}$ construite avec une erreur de mesure sur la position des antennes, c'est à dire 
@@
 	\end{equation}
+
+	On peut de même évaluer le différentiel sur les covariances :
+	\begin{equation}
+		\Delta P_{k|k} = P_{k|k} - P_{k|k}^c = \left(K^c\left(\Hc\right) - K\Ht\right)P_{k|k-1}
+	\end{equation}
+
+	De plus, si on suppose une erreur de gain sur les antennes, on peut poser les termes :
+	$$
+		\Ht_{j,q} = (1+\varepsilon_q)H_{j,q}
+	$$
+
+	Où $\varepsilon_q$ suit une loi normale, et désigne une erreur sur le gain de l'antenne $q$.\\
+	On peut de même établir une matrice de correction $\B$ telle que :
+	\begin{equation}
+		H = \Hb
+	\end{equation}
+
+	Les erreurs et gain auront alors la même forme
+	\begin{subequations}
+		\begin{equation}
+			K_k^b = P\left(\Hb\right)^H\left(\left(\Hb\right)P\left(\Hb\right)^H + R\right)^{-1}
+		\end{equation}
+		\begin{equation}
+			\varepsilon_{\widehat{x}} = K\left(\mathbb{1} - \B\right)\odot\Ht\widehat{x}_{k|k-1}
+		\end{equation}
+		\begin{equation}
+			\Delta P_{k|k} = \left(K^b\left(\Hb\right) - K\Ht\right)P_{k|k-1}
+		\end{equation}
+	\end{subequations}
+
+	\subsubsection{Essais expérimentaux}
+
+	On effectue des essais en introduisant une erreur sur la matrice exacte. Plutôt que d'avoir une matrice erronée $\Ht$ connue on aura $H$ exacte connue et :
+	$$
+		\Ht = \mathcal{E}\odot H
+	$$
+	où $\mathcal{E} = \hinv{\C}$ ou $\mathcal{E} = \hinv{\B}$ en fonction de l'erreur que l'on souhaite introduire.
+
+	\paragraph{Erreur de position}
+	On suppose tout d'abord que l'on ne connaît pas exactement la position des antennes du réseau.
+On introduit sur $H$ une matrice de petites rotations définie par la matrice $\mathcal{R} = \hinv{\B}$ plus tôt.
+\\
+	Pour cela, on génère $J^2$ vecteurs $\{\boldsymbol{\varepsilon}_j\}$ et on génère $\mathcal{R}$. 
 \end{document}

### 2022-06-13 Kalman/KalmainV4.m

@@ -14,7 +14,7 @@ Nreal = 1000;
 
 % génération image initiale
 % ajout bruit image
-poids = 0;
+poids = 0.01;
 X_0 = true_image ;
 tX(:,1) = X_0;
 
@@ -24,8 +24,14 @@ P_0 = X_0*X_0';
 
 %ajout erreur sur A (!!!APRES!!! dataGen => sinon c'est pas une erreur)
 % H_err = abs(H .*( poids*exp(1j*randn(size(H))) ) );
-z_err = z+ randn(size(z))*poids;
-H_err = matF(J,Pix,z_err,lambda,I);
+% z_err = z+ randn(size(z))*poids;
+% H_err = matF(J,Pix,z_err,lambda,I);
+
+H_err = zeros(size(H));
+eps = randn(size(z))*poids;
+for q=1:Pix
+    H_err(: ,q) = exp(-1i*2*pi*eps(:,q)*I(q,:));
+end % .* aussi non ?
 
 %%%%%%%%%%%% Kalman
 X=-1;

### 2022-06-13 Kalman/KalmainV4.m
çamarchpa .__.

@@ -14,7 +14,7 @@ Nreal = 1000;
 
 % génération image initiale
 % ajout bruit image
-poids = 0.01;
+poids = 0.1;
 X_0 = true_image ;
 tX(:,1) = X_0;
 
@@ -24,14 +24,9 @@ P_0 = X_0*X_0';
 
 %ajout erreur sur A (!!!APRES!!! dataGen => sinon c'est pas une erreur)
 % H_err = abs(H .*( poids*exp(1j*randn(size(H))) ) );
-% z_err = z+ randn(size(z))*poids;
-% H_err = matF(J,Pix,z_err,lambda,I);
+z_err = z+ randn(size(z))*poids;
+H_err = matF(J,Pix,z_err,lambda,I);
 
-H_err = zeros(size(H));
-eps = randn(size(z))*poids;
-for q=1:Pix
-    H_err(: ,q) = exp(-1i*2*pi*eps(:,q)*I(q,:));
-end % .* aussi non ?
 
 %%%%%%%%%%%% Kalman
 X=-1;
@@ -39,15 +34,15 @@ if mode=='CPU'
     X = Kalman_CPU_V2(A,H_err,X_0,P_0,nY,nR,nQ,Pix,N);
 end
 if mode=='GPU'
-    X = Kalman_GPU_V4(A,H,X_0,P_0,nY,nR,nQ,Pix,N);
+    X = Kalman_GPU_V5(A,H_err,X_0,P_0,nY,nR,nQ,Pix,N);
 end
 
 Psr = zeros(N,1);
 for k=1:N
-    %     Psr(k) = psnr(abs(X(:,k)),tX(:,k));
-    Psr(k) = ssim(abs(X(:,k)/max(abs(X(:,k)))),tX(:,k));
+        Psr(k) = psnr(abs(X(:,k)),tX(:,k));
+%     Psr(k) = ssim(abs(X(:,k)/max(abs(X(:,k)))),tX(:,k));
 end
-% figure();plot(Psr);title('PSNR = f(it)')
+figure();plot(Psr);title('PSNR = f(it)')
 mean(Psr(6:end))
 
 %{

### 2022-06-13 Kalman/Kalman_GPU_V5.m

@@ -1,0 +1,37 @@
+function X = Kalman_GPU_V5(A,H,X_0,P_0,nY,nR,nQ,Pix,n)
+%KALMAN_GPU Estime via un filtre de Kalman la sortie X-out depuis X_in
+%   variables:
+%   - H : observation (J^2 x Pix) -> matrice de transformation FFT non
+%           uniforme
+%   - A : matrice d'évolution (Pix x Pix)
+%   - X_in : image initiale (Pix x 1)
+%   - R : matrice de covariance du bruit d'observation
+%   - Q : matrice de covariance de l'erreur sur le modèle d'évolution
+%   - I : nombre itérations
+
+X = zeros(Pix,n);
+P = zeros(Pix,Pix,n);
+
+%%% INITIALISATION
+X(:,1) = X_0;
+P(:,:,1) = P_0;
+% figure();
+for k=2:n
+    fprintf('itération : %u \n',k);
+    %%% PREDICTION
+    Xp = A*X(:,k-1) + transpose(mean(nQ(:,:,k-1)));
+    Pp = A*P(:,:,k-1)*(A.') + nQ(:,:,k);
+    yp = H*Xp + mean(nR(:,k));
+    %%% KALMAN GAIN
+    S = gpuArray(H*Pp*(H.') + nR(:,k));
+    T = gpuArray(Pp*(H.'));
+    K = T*pinv(S);
+    K = gather(K);
+    clear S; clear T;
+    %%% ESTIMATION
+    X(:,k) = Xp + K*(nY(:,k)-yp);
+    P(:,:,k) = Pp - K*H*Pp;
+%     subplot(n-1,1,k-1);
+%     imshow(abs(K));colorbar;axis image;colormap("parula");title(mean(K, 'all'))
+end
+end

### 2022-06-13 Kalman/dispKalman_t.m

@@ -4,7 +4,7 @@ function  dispKalman_t(X,tX,N,Mx,My)
 figure();
 for k=1:N
     subplot(2,N,k);imshow(y_to_im(abs(tX(:,k)),Mx,My));colorbar;axis image;colormap("parula")
-    subplot(2,N,k+N);imshow(y_to_im(abs(X(:,k)/max(abs(X(:,k)))),Mx,My));colorbar;axis image;colormap("parula");title(ssim(abs(X(:,k)/max(abs(X(:,k)))),tX(:,k)))
+    subplot(2,N,k+N);imshow(y_to_im(abs(X(:,k)/max(abs(X(:,k)))),Mx,My));colorbar;axis image;colormap("parula");title(psnr(abs(X(:,k)/max(abs(X(:,k)))),tX(:,k)))
 end
 end
 

### 2022-06-13 main_2.m

@@ -10,11 +10,11 @@ c= 3e8;
 f = 1e9;
 lambda = c/f;
 dist = 10; % J_b antennes espacées de dist 
-N = 7;
+N = 12;
 SNR = 10;
 iMEM = 1; % nombre itérations MEM 
 RATIO = 10e14; % ratio erreur Kalman
-mode = 'CPU';
+mode = 'GPU';
 % A_ev matrice d'évolution définie plus loin (need Pix ligne 69)
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@ -39,4 +39,4 @@ A_ev = eye(Pix);
 
 %%% Kalman direct en fait mdr
 [X,tX] = KalmainV4(vadapted,J,z,I,N,f,c,lambda,Mx,My,iMEM,SNR,A_ev,RATIO,mode);
-dispKalman_t(X,tX,N,Mx,My);
+% dispKalman_t(X,tX,N,Mx,My);

### 2022-06-14 Kalman/KalmainV4.m
dfghj

@@ -14,7 +14,7 @@ Nreal = 1000;
 
 % génération image initiale
 % ajout bruit image
-poids = 0.1;
+poids = 0.01;
 X_0 = true_image ;
 tX(:,1) = X_0;
 
@@ -39,8 +39,8 @@ end
 
 Psr = zeros(N,1);
 for k=1:N
-        Psr(k) = psnr(abs(X(:,k)),tX(:,k));
-%     Psr(k) = ssim(abs(X(:,k)/max(abs(X(:,k)))),tX(:,k));
+%         Psr(k) = psnr(abs(X(:,k)),tX(:,k));
+    Psr(k) = psnr(abs(X(:,k)/max(abs(X(:,k)))),tX(:,k));
 end
 figure();plot(Psr);title('PSNR = f(it)')
 mean(Psr(6:end))

### 2022-06-14 main_2.m

@@ -14,7 +14,7 @@ N = 12;
 SNR = 10;
 iMEM = 1; % nombre itérations MEM 
 RATIO = 10e14; % ratio erreur Kalman
-mode = 'GPU';
+mode = 'CPU';
 % A_ev matrice d'évolution définie plus loin (need Pix ligne 69)
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@ -39,4 +39,4 @@ A_ev = eye(Pix);
 
 %%% Kalman direct en fait mdr
 [X,tX] = KalmainV4(vadapted,J,z,I,N,f,c,lambda,Mx,My,iMEM,SNR,A_ev,RATIO,mode);
-% dispKalman_t(X,tX,N,Mx,My);
+dispKalman_t(X,tX,N,Mx,My);

### 2022-06-14 rendu/résultats_brouillon/r_brouillon.tex
çamarchpa .__.
14-06

@@
+% !TeX document-id = {2385fae9-28e6-4c73-97cb-b86e8670f6e7}
 \documentclass[titlepage]{article}
@@
 \usepackage{stmaryrd}
+\usepackage{minted}
 \DeclareMathOperator{\e}{e}
@@
 
-\newcommand{\C}{\mathcal{C}}
-\newcommand{\Ht}{\widetilde{H}}
-\newcommand{\Hc}{\C\odot\Ht}
-\newcommand{\B}{\mathcal{B}}
-\newcommand{\Hb}{\B\odot\Ht}
-\newcommand{\hinv}[1]{#1^{\circ-1}}
+\newcommand{\C}{\mathcal{C}} % mat correction
+\newcommand{\Ht}{\widetilde{H}} % H err
+\newcommand{\Hc}{\C\odot\Ht} % C o H
+\newcommand{\B}{\mathcal{B}} % mat correction gain
+\newcommand{\Hb}{\B\odot\Ht} % B o H
+\newcommand{\hinv}[1]{#1^{\circ-1}} % inverse de hadamard
+\newcommand{\fnorm}[1]{|\vert#1|\vert_{F}} % norme de Frobenius
 
@@
 		\forall A \in \left(\mathbb{C}\backslash\{0\}\right)^{m\times n},\quad \hinv{A}\in\left(\mathbb{C}\backslash\{0\}\right)^{m\times n} \quad\text{et}\quad \forall (i,j)\in\llbracket1;m\rrbracket\times\llbracket1;n\rrbracket,\, \left(\hinv{A}\right)_{i,j} = \left(A_{i,j}\right)^{-1}
+	$$
+
+	Ainsi que la notation $\oslash$ désignant la division de Hadamard :
+		$$
+	\forall (A,B)\in\mathbb{C}^{m\times n}\times\left(\mathbb{C}\backslash0\right)^{m\times n}, \quad A\oslash B\in\mathbb{C}^{m\times n} \quad\text{et}\quad \forall (i,j)\in\llbracket1;m\rrbracket\times\llbracket1;n\rrbracket,\, (A\oslash B)_{i,j} = \frac{A_{i,j}}{B_{i,j}}
+	$$
+
+	\subsection{Norme de Frobenius}
+
+	On dénote $\fnorm{\bullet}$ la norme de Frobenius :
+	$$
+		A\in\mathcal{M}_{m,n}(\mathbb{K}) \quad \fnorm{A} := \sqrt{\tr{AA^H}} = \sqrt{\;\;\;\sum_{\mathclap{
+					\substack{
+						1\le i\le m \\1\le j\le n}
+					}
+				}\abs{A_{ij}}^2}
 	$$
@@
 	\begin{equation}
-		\varepsilon_K = K^c - K
+		\varepsilon_K = K - K^c 
 	\end{equation}
@@
 		&= \left(K-K^c\right)y_k - \left(K\Ht - K^c(\Hc)\right)\widehat{x}_{k|k-1} \\
-		&= \varepsilon_K\left[y_k - \left(\Hc\right)\widehat{x}_{k|k-1}\right] + K\left(\mathbb{1}-\C\right)\odot\Ht\widehat{x}_{k|k-1}
+		&= \varepsilon_K\left[y_k - \left(\Hc\right)\widehat{x}_{k|k-1}\right] + K\left(\mathbb{1}-\C\right)\odot\Ht\widehat{x}_{k|k-1} \\
+		&= \varepsilon_K(y_k - (\Hc)\left(\varepsilon_{pred} + x_k\right)) + K\left(\mathbb{1}-\C\right)\odot\Ht\widehat{x}_{k|k-1} \\
+		&= \varepsilon_K\left[\left(y_k - (\Hc)x_k\right) - (\Hc)\varepsilon_{pred}\right] + K\left(\mathbb{1}-\C\right)\odot\Ht\widehat{x}_{k|k-1} \\
+		&= \varepsilon_K\left(\varepsilon_{mes} - (\Hc)\varepsilon_{pred}\right) + K\left(\mathbb{1}-\C\right)\odot\Ht\widehat{x}_{k|k-1}
 	\end{align*}
@@
 	Où $\mathbb{1}$ désigne l'élément neutre du produit de Hadamard, la matrice dont tous les termes valent 1.\\
-	Donc :
-	\begin{equation}
-		\varepsilon_{\widehat{x}} = \varepsilon_K\varepsilon_{mes} + K\left(\mathbb{1} - \mathcal{C}\right)\odot\Ht\widehat{x}_{k|k-1}
-	\end{equation}
 
@@
 	$$
-		y_k = \Hc x_k \implies \varepsilon_{mes} = 0
+		y_k = (\Hc) x_k \implies \varepsilon_{mes} = 0
 	$$
+	On ne suppose également pas d'erreur sur la matrice d'évolution $A$ :
+	$$
+		\varepsilon_{pred} = 0
+	$$
+
 	D'où :
@@
 	$$
-		\Ht_{j,q} = (1+\varepsilon_q)H_{j,q}
+		\Ht_{j,q} = (1+\varepsilon_{j,q})H_{j,q}
 	$$
 
-	Où $\varepsilon_q$ suit une loi normale, et désigne une erreur sur le gain de l'antenne $q$.\\
+	Où $\varepsilon_q$ suit une loi normale, et désigne une erreur sur le gain de l'antenne $j$ dans la direction $q$.\\
 	On peut de même établir une matrice de correction $\B$ telle que :
@@
 	On suppose tout d'abord que l'on ne connaît pas exactement la position des antennes du réseau.
-On introduit sur $H$ une matrice de petites rotations définie par la matrice $\mathcal{R} = \hinv{\B}$ plus tôt.
+On introduit sur $H$ une matrice de petites rotations définie par la matrice $\mathcal{R} = \hinv{\C}$ plus tôt.
 \\
-	Pour cela, on génère $J^2$ vecteurs $\{\boldsymbol{\varepsilon}_j\}$ et on génère $\mathcal{R}$. 
+	Pour cela, on génère des erreurs directement sur les positions des antennes, puis on peut obtenir $\C$ avec la connaissance de $H$ et $\Ht$ :
+	\begin{equation}
+		\C = H \oslash \Ht
+	\end{equation}
 \end{document}

### 2022-06-14 Kalman/KalmainV4.m

@@ -14,7 +14,7 @@ Nreal = 1000;
 
 % génération image initiale
 % ajout bruit image
-poids = 0.01;
+poids = 0.1;
 X_0 = true_image ;
 tX(:,1) = X_0;
 
@@ -26,6 +26,8 @@ P_0 = X_0*X_0';
 % H_err = abs(H .*( poids*exp(1j*randn(size(H))) ) );
 z_err = z+ randn(size(z))*poids;
 H_err = matF(J,Pix,z_err,lambda,I);
+
+C = H ./ H_err;
 
 
 %%%%%%%%%%%% Kalman

### 2022-06-14 main_2.m

@@ -9,7 +9,7 @@ J_b = 6; % On suppose le réseau d'antennes carré de côté
 c= 3e8;
 f = 1e9;
 lambda = c/f;
-dist = 10; % J_b antennes espacées de dist 
+dist = 10*lambda/2.1; % J_b antennes espacées de dist 
 N = 12;
 SNR = 10;
 iMEM = 1; % nombre itérations MEM 

### 2022-06-14 sim/matF.m

@@ -12,7 +12,7 @@ end
 dz = reshape(dz,J^2,2);
 
 for q=1:Q
-    F(: ,q) = exp(-1i*2*pi*(dz(:,1)*I(q,1)+dz(:,2)*I(q,2)));
+    F(: ,q) = exp(-1i*2*pi/lambda*(dz(:,1)*I(q,1)+dz(:,2)*I(q,2)));
 end % .* aussi non ?
 end
 

### 2022-06-20 Kalman/KalmainV4.m
!!! rnA pas trop grand A (ifft pix*pix)

@@ -14,7 +14,7 @@ Nreal = 1000;
 
 % génération image initiale
 % ajout bruit image
-poids = 0.1;
+poids = 0;
 X_0 = true_image ;
 tX(:,1) = X_0;
 

### 2022-06-20 img_t/rnA.m

@@ -1,0 +1,27 @@
+function Ad = rnA(A,F,p)
+%RNA déforme la matrice d'entrée A
+
+Ad=A; % default si mode incohérent
+
+pente = 0.1;
+GausSize = 3; % taille matrice gaussienne (eg 3 = 3x3)
+GaussCentr = 0.9; % poids centre 
+
+% génération champ de distortion
+
+
+N =size(A); % size in pixels of image
+       % frequency-filter width
+
+[X,Y] = ndgrid(1:N(1),1:N(2));
+i = min(X-1,N(1)-X+1);
+j = min(Y-1,N(2)-Y+1);
+H = exp(-.5*(i.^2+j.^2)/F^2);
+Z = real(ifft2(H.*fft2(randn(N))));
+
+surf(Z);
+
+Ad = p*Z + A;
+
+end
+

### 2022-06-20 main_2.m

@@ -3,14 +3,14 @@ close all
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 % PARAMETRES
-Mx = 5; %My auto (ratio)
+Mx = 20; %My auto (ratio)
 J_b = 6; % On suppose le réseau d'antennes carré de côté
 % max 11 GPU atm
 c= 3e8;
 f = 1e9;
 lambda = c/f;
 dist = 10*lambda/2.1; % J_b antennes espacées de dist 
-N = 12;
+N = 3; % itérations Kalman
 SNR = 10;
 iMEM = 1; % nombre itérations MEM 
 RATIO = 10e14; % ratio erreur Kalman
@@ -19,7 +19,7 @@ mode = 'CPU';
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
 %%% choix / gen image
-img = imread('disques.jpg');
+img = imread('nasa.jpg');
 adapted = adaptImg(img, Mx); 
 vadapted = vectorize(adapted);
 
@@ -37,6 +37,15 @@ A_ev = eye(Pix);
 % A_ev = zeros(Pix);
 % A_ev(7,1) = 1; A_ev(4,2)=1; A_ev(1,3)=1; A_ev(8,4)=1;A_ev(5,5)=1;A_ev(2,6)=1;A_ev(9,7)=1;A_ev(6,8)=1;A_ev(3,9)=1;
 
-%%% Kalman direct en fait mdr
-[X,tX] = KalmainV4(vadapted,J,z,I,N,f,c,lambda,Mx,My,iMEM,SNR,A_ev,RATIO,mode);
-dispKalman_t(X,tX,N,Mx,My);
+poids = 0.1;
+Fw = 1.5;
+A_err = rnA(A_ev,Fw,poids);
+
+% test image
+evd = A_err*vadapted;
+figure();imshow(y_to_im(abs(vadapted),Mx,My));colorbar;axis image;colormap("parula")
+figure();imshow(y_to_im(abs(evd),Mx,My));colorbar;axis image;colormap("parula");title("ev_err")
+
+%%% Kalman
+% [X,tX] = KalmainV4(vadapted,J,z,I,N,f,c,lambda,Mx,My,iMEM,SNR,A_err,RATIO,mode);
+% dispKalman_t(X,tX,N,Mx,My);

### 2022-06-21 rendu/résultats_brouillon/r_brouillon.tex
update rapide
up21-06

@@
 	$$
-		A\in\mathcal{M}_{m,n}(\mathbb{K}) \quad \fnorm{A} := \sqrt{\tr{AA^H}} = \sqrt{\;\;\;\sum_{\mathclap{
+		A\in\mathcal{M}_{m,n}(\mathbb{K}) \quad \fnorm{A} := \sqrt{\tr{AA^H}} = \sqrt{
+				\;\;\;\sum_{
+					\mathclap{
 						\substack{
-						1\le i\le m \\1\le j\le n}
+							1\le i\le m \\1\le j\le n
+						}
 					}
 				}\abs{A_{ij}}^2}
+	$$
+
+	Qui induit sur $\mathcal{M}_{m,n}(\mathbb{K})$ la distance $d$ :
+	$$
+		\forall(A,B)\in\left(\mathcal{M}_{m,n}(\mathbb{K})\right)^2\quad d(A,B) = \frac{\fnorm{A-B}}{mn}
 	$$
@@
 	On suppose une erreur sur le modèle d'évolution \emph{dans le filtre de Kalman}, afin de simuler un modèle théorique imprécis. On construit les données simulées avec un modèle sans erreur.\\
-	\begin{minted}{matlab}
-poids = 0.2;
-A = A + poids*randn(size(A));
-	\end{minted}
 
-	On introduit une erreur gaussienne
+	\subsubsection{Calculs théoriques}
+
+	On suppose qu'on a à disposition un modèle d'évolution déformé 
+	\begin{equation}
+		A_d = A\ast\varepsilon
+	\end{equation}
+
+	où $\ast$ désigne l'opération de convolution. \\
+	On obtient le nouveau modèle :
+	\begin{equation}
+		\widehat{x}_{k|k-1} = \left(A\ast\varepsilon\right)\widehat{x}_{k-1|k-1}
+	\end{equation}
+
+	La covariance de l'erreur de prédiction devient :
+	\begin{align*}
+		P_{k|k-1} &= E\left[\left(A_d\widehat{x}_{k-1|k-1} - x_k\right)\left(A_d\widehat{x}_{k-1|k-1} - x_k\right)^H\right]\\
+		&= E\left\{
+			\left[
+				A_d(\widehat{x}_{k-1|k-1} - x_{k-1}) + (A_dx_{k-1} - x_k)
+			\right]\left[
+			A_d(\widehat{x}_{k-1|k-1} - x_{k-1}) + (A_dx_{k-1} - x_k)
+			\right]^H
+		\right\} \\
+		&= A_dP_{k-1|k-1}Ad^T + Q_d + 0
+	\end{align*}
+
+	Le dernier terme s'explicant par la décorrélation des erreurs sur le modèle d'évolution et d'erreur de prédiction. On obtient finalement le modèle :
+	\begin{equation}
+		P_{k|k-1} = \left(A\ast\varepsilon\right)P_{k-1|k-1}\left(A\ast\varepsilon\right)^T + Q_d
+	\end{equation}
 
@@
 	\end{equation}
+
+	Cette division est possible car tous les termes sont de norme 1 ou proche de 1 dans le cas de petites variations.
 \end{document}

### 2022-06-21 Kalman/KalmainV4.m

@@ -14,7 +14,7 @@ Nreal = 1000;
 
 % génération image initiale
 % ajout bruit image
-poids = 0;
+poids = 0.3;
 X_0 = true_image ;
 tX(:,1) = X_0;
 
@@ -42,10 +42,10 @@ end
 Psr = zeros(N,1);
 for k=1:N
 %         Psr(k) = psnr(abs(X(:,k)),tX(:,k));
-    Psr(k) = psnr(abs(X(:,k)/max(abs(X(:,k)))),tX(:,k));
+    Psr(k) = fro(abs(X(:,k)/max(abs(X(:,k))))-tX(:,k));
 end
 figure();plot(Psr);title('PSNR = f(it)')
-mean(Psr(6:end))
+mean(Psr(4:end))
 
 %{
 z_err = z+ randn(size(z))*0.01;

### 2022-06-21 Kalman/Kalman_CPU_V3.m

@@ -1,0 +1,33 @@
+function [X,K] = Kalman_CPU_V3(A,H,X_0,P_0,nY,nR,nQ,Pix,n)
+%KALMAN_CPU Estime via un filtre de Kalman la sortie X-out depuis X_in
+%   variables:
+%   - H : observation (J^2 x Pix) -> matrice de transformation FFT non
+%           uniforme
+%   - A : matrice d'évolution (Pix x Pix)
+%   - X_in : image initiale (Pix x 1)
+%   - R : matrice de covariance du bruit d'observation
+%   - Q : matrice de covariance de l'erreur sur le modèle d'évolution
+%   - I : nombre itérations
+
+X = zeros(Pix,n);
+P = zeros(Pix,Pix,n);
+
+%%% INITIALISATION
+X(:,1) = X_0;
+P(:,:,1) = P_0;
+% figure();
+for k=2:n
+    fprintf('itération : %u \n',k);
+    %%% PREDICTION
+    Xp = A*X(:,k-1) + transpose(mean(nQ(:,:,k-1)));
+    Pp = A*P(:,:,k-1)*(A.') + nQ(:,:,k);
+    yp = H*Xp + mean(nR(:,k));
+    %%% KALMAN GAIN
+    K = Pp*(H.')*pinv(H*Pp*(H.') + nR(:,k));
+    %%% ESTIMATION
+    X(:,k) = Xp + K*(nY(:,k)-yp);
+    P(:,:,k) = Pp - K*H*Pp;
+%     subplot(n-1,1,k-1);
+%     imshow(abs(K));colorbar;axis image;colormap("parula");title(mean(K, 'all'))
+end
+end

### 2022-06-21 Kalman/dispKalman_t.m

@@ -4,7 +4,7 @@ function  dispKalman_t(X,tX,N,Mx,My)
 figure();
 for k=1:N
     subplot(2,N,k);imshow(y_to_im(abs(tX(:,k)),Mx,My));colorbar;axis image;colormap("parula")
-    subplot(2,N,k+N);imshow(y_to_im(abs(X(:,k)/max(abs(X(:,k)))),Mx,My));colorbar;axis image;colormap("parula");title(psnr(abs(X(:,k)/max(abs(X(:,k)))),tX(:,k)))
+    subplot(2,N,k+N);imshow(y_to_im(abs(X(:,k)/max(abs(X(:,k)))),Mx,My));colorbar;axis image;colormap("parula");title(fro(abs(X(:,k)/max(abs(X(:,k)))) - tX(:,k)))
 end
 end
 

### 2022-06-21 main_2.m

@@ -3,14 +3,14 @@ close all
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 % PARAMETRES
-Mx = 20; %My auto (ratio)
+Mx = 5; %My auto (ratio)
 J_b = 6; % On suppose le réseau d'antennes carré de côté
 % max 11 GPU atm
 c= 3e8;
 f = 1e9;
 lambda = c/f;
 dist = 10*lambda/2.1; % J_b antennes espacées de dist 
-N = 3; % itérations Kalman
+N = 6; % itérations Kalman
 SNR = 10;
 iMEM = 1; % nombre itérations MEM 
 RATIO = 10e14; % ratio erreur Kalman
@@ -39,13 +39,13 @@ A_ev = eye(Pix);
 
 poids = 0.1;
 Fw = 1.5;
-A_err = rnA(A_ev,Fw,poids);
+% A_err = rnA(A_ev,Fw,poids);
 
 % test image
-evd = A_err*vadapted;
-figure();imshow(y_to_im(abs(vadapted),Mx,My));colorbar;axis image;colormap("parula")
-figure();imshow(y_to_im(abs(evd),Mx,My));colorbar;axis image;colormap("parula");title("ev_err")
+% % evd = A_err*vadapted;
+% figure();imshow(y_to_im(abs(vadapted),Mx,My));colorbar;axis image;colormap("parula")
+% figure();imshow(y_to_im(abs(evd),Mx,My));colorbar;axis image;colormap("parula");title("ev_err")
 
 %%% Kalman
-% [X,tX] = KalmainV4(vadapted,J,z,I,N,f,c,lambda,Mx,My,iMEM,SNR,A_err,RATIO,mode);
-% dispKalman_t(X,tX,N,Mx,My);
+[X,tX] = KalmainV4(vadapted,J,z,I,N,f,c,lambda,Mx,My,iMEM,SNR,A_ev,RATIO,mode);
+dispKalman_t(X,tX,N,Mx,My);

### 2022-06-21 main_loop.m

@@ -1,0 +1,3 @@
+for i=1:5
+    main_2;
+end

### 2022-06-21 utils/fro.m

@@ -1,0 +1,5 @@
+function fr = fro(A)
+%FRO Calcul norme frobenius
+fr = trace(A*(A'))/numel(A);
+end
+

### 2022-06-22 Kalman/Kalman_GPU_V5.m
nbvc

@@ -17,7 +17,7 @@ X(:,1) = X_0;
 P(:,:,1) = P_0;
 % figure();
 for k=2:n
-    fprintf('itération : %u \n',k);
+%     fprintf('itération : %u \n',k);
     %%% PREDICTION
     Xp = A*X(:,k-1) + transpose(mean(nQ(:,:,k-1)));
     Pp = A*P(:,:,k-1)*(A.') + nQ(:,:,k);

### 2022-06-23 main_loop.m
ca marche <3<3<3

@@ -1,3 +1,4 @@
-for i=1:5
+for i=1:6
+    fprintf('itération : %u ---',i);
     main_2;
 end

### 2022-06-24 rendu/résultats_brouillon/r_brouillon.tex
update_midi
Zzz

@@
 \usepackage[T1]{fontenc}
+\usepackage{svg}
 
@@
 	Cette division est possible car tous les termes sont de norme 1 ou proche de 1 dans le cas de petites variations.
+
+	On effectue 6 réalisations de Kalman sur 8 instants temporels en introduisant une erreur sur la position des antennes.
+On effectue cette opération pour plusieurs erreurs et on peut observer les erreurs $\fnorm{\varepsilon_K}$ et $\fnorm{\varepsilon_{\widehat{x}}}$ :
+
+	\begin{figure}[H]
+		\centering
+		\begin{subfigure}{.5\textwidth}
+			\centering
+			\includesvg[width=.9\linewidth]{Perfs/e_K.svg}
+		\end{subfigure}%
+		\begin{subfigure}{.5\textwidth}
+			\centering
+			\includesvg[width=.9\linewidth]{Perfs/e_x.svg}
+		\end{subfigure}
+		\caption{Erreur induite sur les matrices par un erreur de position}
+	\end{figure}
+
+	On peut également tracer l'erreur commise sur la reconstruction en fonction de l'erreur sur la matrice K :
+
+	\begin{figure}[H]
+		\centering
+		\includesvg[width=0.7\linewidth]{Perfs/x_fK}
+		\caption{Erreur de reconstruction en fonction de l'erreur sur la matrice K}
+	\end{figure}
+
 \end{document}

### 2022-06-24 Kalman/KalmainV4.m

@@ -47,7 +47,7 @@ for k=1:N
 %     Psr(k) = fro(K_err(:,:,k)-K(:,:,k));
 end
 % figure();plot(Psr);title('PSNR = f(it)')
-% fprintf('||e_K|| = %e \n',mean(Psr(4:end)));
+fprintf('||e_X|| = %e ---- ||e_X||th = %e \n',mean(Psr(4:end)),mean(e_x(4:end)));
 
 
 figure();plot(e_x);title("Ev e_x")

### 2022-06-24 main_2.m

@@ -10,7 +10,7 @@ c= 3e8;
 f = 1e9;
 lambda = c/f;
 dist = 10*lambda/2.1; % J_b antennes espacées de dist 
-N = 6; % itérations Kalman
+N = 7; % itérations Kalman
 SNR = 10;
 iMEM = 1; % nombre itérations MEM 
 RATIO = 10e14; % ratio erreur Kalman
@@ -48,4 +48,4 @@ Fw = 1.5;
 
 %%% Kalman
 [X,tX] = KalmainV4(vadapted,J,z,I,N,f,c,lambda,Mx,My,iMEM,SNR,A_ev,RATIO,mode);
-dispKalman_t(X,tX,N,Mx,My);
+% dispKalman_t(X,tX,N,Mx,My);

### 2022-06-27 rendu/résultats_brouillon/r_brouillon.tex
ajout d2 + clean d1 + figures

@@
 
-	\subsection{Norme de Frobenius}
+	\subsection{Norme de Frobenius et distances}
 
@@
 
-	Qui induit sur $\mathcal{M}_{m,n}(\mathbb{K})$ la distance $d$ :
+	Qui induit sur $\mathcal{M}_{m,n}(\mathbb{D})$ la distance normalisée $d_1$ :
 	$$
-		\forall(A,B)\in\left(\mathcal{M}_{m,n}(\mathbb{K})\right)^2\quad d(A,B) = \frac{\fnorm{A-B}}{mn}
+		\forall(A,B)\in\left(\mathcal{M}_{m,n}(\mathbb{D})\right)^2\quad d_1(A,B) = \frac{\fnorm{A-B}}{mn}
+	$$
+
+	Où $\mathbb{D}$ désigne le disque unité fermé.\\
+
+	Ainsi que la distance $d_2$ :
+	$$
+		\forall(A,B)\in\left(\mathcal{M}_{m,n}(\mathbb{K})\right)^2\quad d_2(A,B) = \sum_{
+			\mathclap{
+				\substack{
+					1\le i\le m \\1\le j\le n
+				}
+			}
+		}\abs{A_{ij} - B_{i,j}}
 	$$
@@
 
-	\subsection{Erreur sur le masque UV}
+	\subsection{Erreur sur le masque UV - Position}
 
@@
 	D'où :
-	\begin{equation}
+	\begin{equation}\label{eq:approx}
 		\varepsilon_{\widehat{x}} = K \left(\mathbb{1} - \C\right)\odot\Ht\widehat{x}_{k|k-1}
@@
 	On effectue 6 réalisations de Kalman sur 8 instants temporels en introduisant une erreur sur la position des antennes.
-On effectue cette opération pour plusieurs erreurs et on peut observer les erreurs $\fnorm{\varepsilon_K}$ et $\fnorm{\varepsilon_{\widehat{x}}}$ :
+On effectue cette opération pour plusieurs erreurs et on peut observer les erreurs $\varepsilon_K$ et $\varepsilon_{\widehat{x}}$ calculées par la distance $d_1$, normalisée sur le disque unité :
 
@@
 		\centering
-		\includesvg[width=0.7\linewidth]{Perfs/x_fK}
+		\includesvg[width=0.6\linewidth]{Perfs/x_fK}
 		\caption{Erreur de reconstruction en fonction de l'erreur sur la matrice K}
 	\end{figure}
+
+	On peut également vérifier la précision de l'approximation pour obtenir l'équation (\ref{eq:approx}) :
+
+	\begin{figure}[H]
+		\centering
+		\includesvg[width=.7\linewidth]{Perfs/graph_approx.svg}
+		\caption{Essai sur une réalisation de 10 instants}
+	\end{figure}
+
+	On peut également tracer la robustesse à l'erreur de cette approximation :
+
+	\begin{figure}[H]
+		\centering
+		\includesvg[width=.6\linewidth]{Perfs/approx.svg}
+		\caption{Erreur de reconstruction finale en fonction de l'erreur en position}
+	\end{figure}
+
+	\subsection{Erreur sur le masque UV - direction}
+
+	En considérant une erreur sur la position des antennes :
+	\begin{equation}
+		\forall q\in \llbracket 1;Q\rrbracket,\; \widetilde{\boldsymbol{I}_q} = \boldsymbol{I}_q + \boldsymbol{\varepsilon}_q
+	\end{equation}
+
+	où $Q$ désigne le nombre de directions observées, on obtient un calcul extrêmement similaire à l'erreur sur les positions d'antennes.
+On peut donc effectuer de même des essais afin de vérifier les performances en présence d'une erreur.
+\\
+	On effectue les essais sur 35 directions (quadrillage $7\times5$) et 49 antennes (réseau carré $7\times7$).
+On quantifiera d'abord l'erreur en présence d'une direction faussée, puis de 5 et enfin toutes.
+\\
+
+	\paragraph{Erreur sur une unique direction :} 
+	On utilisera dans cette partie la distance $d_2$, afin d'évaluer l'impact de la direction sur laquelle est faite l'erreur sur l'erreur de reconstruction totale, grâce à l'additivité de la distance sur les termes de la matrice évaluée.
+	% a reformuler
 

### 2022-06-27 Kalman/KalmainV4.m

@@ -24,8 +24,10 @@ P_0 = X_0*X_0';
 
 %ajout erreur sur A (!!!APRES!!! dataGen => sinon c'est pas une erreur)
 % H_err = abs(H .*( poids*exp(1j*randn(size(H))) ) );
-z_err = z+ randn(size(z))*poids;
-H_err = matF(J,Pix,z_err,lambda,I);
+% z_err = z+ randn(size(z))*poids;
+I_err = I;
+I_err(4,:) = I_err(4,:) + randn(size(I_err(4,:)))*poids;
+H_err = matF(J,Pix,z,lambda,I_err);
 
 C = H ./ H_err;
 
@@ -33,27 +35,33 @@ C = H ./ H_err;
 %%%%%%%%%%%% Kalman
 X_err=-1;
 if mode=='CPU'
-    [X_err,e_x] = Kalman_CPU_V3(A,H_err,X_0,P_0,nY,nR,nQ,Pix,N,C);
-%     [~,K] = Kalman_CPU_V3(A,H,X_0,P_0,nY,nR,nQ,Pix,N);
+    [X_err,K_err] = Kalman_CPU_V3(A,H_err,X_0,P_0,nY,nR,nQ,Pix,N,C);
+    [~,K] = Kalman_CPU_V3(A,H,X_0,P_0,nY,nR,nQ,Pix,N);
 end
 if mode=='GPU'
     X_err = Kalman_GPU_V5(A,H_err,X_0,P_0,nY,nR,nQ,Pix,N);
 end
 
-Psr = zeros(N,1);
+errX = zeros(N,1);
+errK = zeros(N,1);
+errP = zeros(N,1);
 for k=1:N
 %         Psr(k) = psnr(abs(X(:,k)),tX(:,k));
-    Psr(k) = fro(abs(X_err(:,k)/max(abs(X_err(:,k))))-tX(:,k));
-%     Psr(k) = fro(K_err(:,:,k)-K(:,:,k));
+    errX(k) = d_2(abs(X_err(:,k)/max(abs(X_err(:,k))))-tX(:,k));
+    errK(k) = fro(K_err(:,:,k)-K(:,:,k));
+    errP(k) = d_2(abs(X_err(4,k)/max(abs(X_err(:,k))))-tX(4,k));
 end
-% figure();plot(Psr);title('PSNR = f(it)')
-fprintf('||e_X|| = %e ---- ||e_X||th = %e \n',mean(Psr(4:end)),mean(e_x(4:end)));
[…]

### 2022-06-27 Kalman/Kalman_CPU_V3.m

@@ -1,4 +1,4 @@
-function [X,e_x] = Kalman_CPU_V3(A,H,X_0,P_0,nY,nR,nQ,Pix,n,C)
+function [X,out_2] = Kalman_CPU_V3(A,H,X_0,P_0,nY,nR,nQ,Pix,n,C)
 %KALMAN_CPU Estime via un filtre de Kalman la sortie X-out depuis X_in
 %   variables:
 %   - H : observation (J^2 x Pix) -> matrice de transformation FFT non
@@ -11,9 +11,9 @@ function [X,e_x] = Kalman_CPU_V3(A,H,X_0,P_0,nY,nR,nQ,Pix,n,C)
 
 X = zeros(Pix,n);
 P = zeros(Pix,Pix,n);
-% Ks = zeros(Pix,size(H,1),n);
 
-e_x = zeros(n,1);
+Ks = zeros(Pix,size(H,1),n);
+% e_x = zeros(n,1);
 
 %%% INITIALISATION
 X(:,1) = X_0;
@@ -32,8 +32,9 @@ for k=2:n
     P(:,:,k) = Pp - K*H*Pp;
 %     subplot(n-1,1,k-1);
 %     imshow(abs(K));colorbar;axis image;colormap("parula");title(mean(K, 'all'))
-%     Ks(:,:,k) = K; % loggin K
-    e_x(k) = fro(K*( (ones(size(C)) - C) .* H )*Xp);
+    Ks(:,:,k) = K; % loggin K
+%     e_x(k) = fro(K*( (ones(size(C)) - C) .* H )*Xp);
 end
-
+out_2 = Ks;
+% out_2 = e_x;
 end

### 2022-06-27 main_2.m

@@ -3,14 +3,14 @@ close all
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 % PARAMETRES
-Mx = 5; %My auto (ratio)
+Mx = 4; %My auto (ratio)
 J_b = 6; % On suppose le réseau d'antennes carré de côté
 % max 11 GPU atm
 c= 3e8;
 f = 1e9;
 lambda = c/f;
 dist = 10*lambda/2.1; % J_b antennes espacées de dist 
-N = 7; % itérations Kalman
+N = 4; % itérations Kalman
 SNR = 10;
 iMEM = 1; % nombre itérations MEM 
 RATIO = 10e14; % ratio erreur Kalman
@@ -19,7 +19,7 @@ mode = 'CPU';
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
 %%% choix / gen image
-img = imread('nasa.jpg');
+img = imread('bh_pic.jpg');
 adapted = adaptImg(img, Mx); 
 vadapted = vectorize(adapted);
 
@@ -48,4 +48,4 @@ Fw = 1.5;
 
 %%% Kalman
 [X,tX] = KalmainV4(vadapted,J,z,I,N,f,c,lambda,Mx,My,iMEM,SNR,A_ev,RATIO,mode);
-% dispKalman_t(X,tX,N,Mx,My);
+dispKalman_t(X,tX,N,Mx,My);

### 2022-06-27 utils/d_2.m

@@ -1,0 +1,5 @@
+function fr = d_2(A)
+%D2 norme additive
+fr = sum(abs(A),'all');
+end
+

### 2022-06-27 rendu/résultats_brouillon/r_brouillon.tex
graph all

@@
 \\
-	On effectue les essais sur 35 directions (quadrillage $7\times5$) et 49 antennes (réseau carré $7\times7$).
+	On effectue les essais sur 24 directions (quadrillage $4\times6$) et 36 antennes (réseau carré $6\times6$).
 On quantifiera d'abord l'erreur en présence d'une direction faussée, puis de 5 et enfin toutes.
@@
 	% a reformuler
+
+	On obtient :
+
+	\begin{figure}[H]
+		\centering
+		\includesvg[width=.6\linewidth]{Perfs/err_I_1.svg}
+		\caption{Erreurs de reconstruction totale et sur la direction faussée}
+	\end{figure}
+
+	\paragraph{Erreur sur toutes les directions}
+	On utilise la distance $d_1$. On introduit une erreur sur toutes les directions.
+
+	\begin{figure}[H]
+		\centering
+		\includesvg[width=.6\linewidth]{Perfs/err_I_all.svg}
+		\caption{Erreurs de reconstruction}
+	\end{figure}
 

### 2022-06-27 main_2.m

@@ -10,7 +10,7 @@ c= 3e8;
 f = 1e9;
 lambda = c/f;
 dist = 10*lambda/2.1; % J_b antennes espacées de dist 
-N = 4; % itérations Kalman
+N = 6; % itérations Kalman
 SNR = 10;
 iMEM = 1; % nombre itérations MEM 
 RATIO = 10e14; % ratio erreur Kalman
@@ -48,4 +48,4 @@ Fw = 1.5;
 
 %%% Kalman
 [X,tX] = KalmainV4(vadapted,J,z,I,N,f,c,lambda,Mx,My,iMEM,SNR,A_ev,RATIO,mode);
-dispKalman_t(X,tX,N,Mx,My);
+% dispKalman_t(X,tX,N,Mx,My);

### 2022-06-27 main_loop.m

@@ -1,4 +1,4 @@
-for i=1:6
+for i=1:4
     fprintf('itération : %u ---',i);
     main_2;
 end

### 2022-07-11 article/article.tex
azerty

--- 2022-06-24 rendu/résultats_brouillon/r_brouillon.tex
@@
-% !TeX document-id = {2385fae9-28e6-4c73-97cb-b86e8670f6e7}
 \documentclass[titlepage]{article}
@@
 \usepackage[utf8]{inputenc}
-\usepackage[french]{babel}
+\usepackage[english]{babel}
 \usepackage[margin=1in]{geometry}
 \usepackage[T1]{fontenc}
-\usepackage{svg}
 
 \usepackage[colorlinks=true,linkcolor=teal]{hyperref}
-\usepackage{amssymb,amsmath,minted,float,graphicx,textcomp,systeme,listings,physics,mathtools,ifthen,esvect,hyperref,fancyhdr,lastpage,marginnote,chngcntr,fancyvrb,subcaption}
-\usepackage{stmaryrd}
-\usepackage{minted}
+\usepackage{amssymb,amsmath,minted,float,graphicx,textcomp,systeme,listings,physics,mathtools,ifthen,esvect,hyperref,fancyhdr,lastpage,marginnote,chngcntr,fancyvrb,subcaption,svg}
+
 \DeclareMathOperator{\e}{e}
@@
 %%%%%%%%%%%% PARAMETRES
-\newcommand{\UE}{Stage}
-\newcommand{\type}{Résultats}%TD-TP-COURS
+\newcommand{\UE}{4XX}
+\newcommand{\type}{}%TD-TP-COURS
 \newcommand{\nb}{0}%nb 
-\newcommand{\sbt}{Reconstruction dynamique d'images radioastronomiques}%soustitre
+\newcommand{\sbt}{}%soustitre
 \author{Gabriel ROBERT-DAUTUN}
-\date{2022}
+\date{Mars 2022}
 
@@
 
-% permet notes + simples en marge à gauche (ig compteur de question)
+% permet notes + simples en marge à gauche (eg compteur de question)
 \reversemarginpar
@@
 
-\newcommand{\C}{\mathcal{C}} % mat correction
-\newcommand{\Ht}{\widetilde{H}} % H err
-\newcommand{\Hc}{\C\odot\Ht} % C o H
-\newcommand{\B}{\mathcal{B}} % mat correction gain
-\newcommand{\Hb}{\B\odot\Ht} % B o H
-\newcommand{\hinv}[1]{#1^{\circ-1}} % inverse de hadamard
-\newcommand{\fnorm}[1]{|\vert#1|\vert_{F}} % norme de Frobenius
-
 \title{%
-	\UE\, - \type\ifthenelse{\nb > 0}{\nb}{} \\
-	\large \sbt}
+	Radioastronomical image reconstruction using Kalman filter under ionospheric perturbations}
 
@@
 
-	\begin{figure}
-		\centering
-		\includegraphics[width=0.7\linewidth]{logo.PNG}
-		\label{fig:logo}
-	\end{figure}
-	\maketitle
+	\begin{titlepage}
+
+	\huge
+	\textbf{Reconstruction dynamique d'images radioastronomiques de trous noirs}
+
+	\vspace{1.5cm}
+	\LARGE
+	\textbf{Gabriel Robert-Dautun}
+
+	\vfill
+	\large
+	Encadré par Mme Vin
+
+	\vspace{0.8cm}
+
+	\includegraphics[width=0.75\textwidth]{logo.PNG}
+
+	\Large
+	Satie
+	France\\
+	July 2022
+
+	\end{titlepage}
 
@@
 	\newpage
-	\section{Notations}
-	\subsection{Produit de Hadamard}
-
-	On introduit la notation $\odot$ associée au produit de Hadamard, qui effectue la multiplication élément à élément de deux matrices de même taille :
-	$$
-	\forall (A,B)\in\left(\mathbb{C}^{m\times n}\right)^2, \quad A\odot B\in\mathbb{C}^{m\times n} \quad\text{et}\quad \forall (i,j)\in\llbracket1;m\rrbracket\times\llbracket1;n\rrbracket,\, (A\odot B)_{i,j} = A_{i,j}\times B_{i,j}
-	$$
-
-	On introduit également l'inverse de Hadamard, qui inverse élément par élément une matrice
-	$$
-		\forall A \in \left(\mathbb{C}\backslash\{0\}\right)^{m\times n},\quad \hinv{A}\in\left(\mathbb{C}\backslash\{0\}\right)^{m\times n} \quad\text{et}\quad \forall (i,j)\in\llbracket1;m\rrbracket\times\llbracket1;n\rrbracket,\, \left(\hinv{A}\right)_{i,j} = \left(A_{i,j}\right)^{-1}
-	$$
-
-	Ainsi que la notation $\oslash$ désignant la division de Hadamard :
-		$$
-	\forall (A,B)\in\mathbb{C}^{m\times n}\times\left(\mathbb{C}\backslash0\right)^{m\times n}, \quad A\oslash B\in\mathbb{C}^{m\times n} \quad\text{et}\quad \forall (i,j)\in\llbracket1;m\rrbracket\times\llbracket1;n\rrbracket,\, (A\oslash B)_{i,j} = \frac{A_{i,j}}{B_{i,j}}
-	$$
-
-	\subsection{Norme de Frobenius}
-
-	On dénote $\fnorm{\bullet}$ la norme de Frobenius :
-	$$
-		A\in\mathcal{M}_{m,n}(\mathbb{K}) \quad \fnorm{A} := \sqrt{\tr{AA^H}} = \sqrt{
-				\;\;\;\sum_{
-					\mathclap{
-						\substack{
-							1\le i\le m \\1\le j\le n
-						}
-					}
-				}\abs{A_{ij}}^2}
-	$$
-
-	Qui induit sur $\mathcal{M}_{m,n}(\mathbb{K})$ la distance $d$ :
-	$$
-		\forall(A,B)\in\left(\mathcal{M}_{m,n}(\mathbb{K})\right)^2\quad d(A,B) = \frac{\fnorm{A-B}}{mn}
-	$$
-
-	\newpage
-	\section{Influence individuelle de l'erreur sur les matrices}
-	\subsection{Erreur sur l'image de départ}
-	On introduit une erreur gaussienne sur l'image $X_0$. On observe que le fonctionnement du Kalman est bien celui attendu dans ce cas là : l'erreur dans l'image se réduit petit a petit.
-
-	\begin{figure}[H]
-		\centering
-		\includegraphics[width=0.9\linewidth]{Perfs/Illustration.JPG}
-		\caption{En haut : images réelles utilisées pour générer les données. En bas : images reconstruites. Titres : PSNR}
-	\end{figure}
-
-	On utilise une transformation de rotation de l'image de 90°.
-On remarque grâce au PSNR que les images reconstruites s'approchent des images réelles, mais pas parfaitement car le filtre prend en compte l'historique.
-Une erreur dans une image va donc se diffuser aux images suivantes (précédentes si backwards).
-On observe la convergence de reconstruction :
-
-	\begin{figure}[H]
-		\centering
-		\includegraphics[width=0.9\linewidth]{Perfs/conv02.png}
-		\caption{Evolution psnr en fonction du nombre d'itérations}
-	\end{figure}
-
-	On converge en 4 itérations au plus a chaque essai : les erreurs gaussienne d'amplitude 0.2 se diffusent sur 4 itérations environ.
-	Vérifié pour des amplitude de 0.1 a 1. Pour des amplitudes plus grandes (5 donc SNR=0.1) parfois 5 itérations sont nécessaires.
-
-	\subsection{Erreur sur les matrices de covariance}
-
-	Les matrices $\{P_k\}$ et $\{Q_k\}$ servent à calculer le gain de Kalman et donc sélectionner quelle proportion mesure/prédiction choisir pour estimer l'image.
-Si on introduit une erreur \emph{uniquement} sur une de ces matrices, l'image prédite et l'image mesurée seront les mêmes (car on y a pas introduit d'erreur) et donc une erreur sur la proportion n'aura aucun effet.\\
-	Une erreur sur ces matrices n'introduit donc pas d'erreur sur la reconstruction de manière directe. Elle peut néanmoins diffuser une erreur déjà présente.
-
-	\subsection{Erreur sur le modèle d'évolution}
-
-	On suppose une erreur sur le modèle d'évolution \emph{dans le filtre de Kalman}, afin de simuler un modèle théorique imprécis. On construit les données simulées avec un modèle sans erreur.\\
-
-	\subsubsection{Calculs théoriques}
-
-	On suppose qu'on a à disposition un modèle d'évolution déformé 
-	\begin{equation}
-		A_d = A\ast\varepsilon
-	\end{equation}
-
-	où $\ast$ désigne l'opération de convolution. \\
-	On obtient le nouveau modèle :
-	\begin{equation}
-		\widehat{x}_{k|k-1} = \left(A\ast\varepsilon\right)\widehat{x}_{k-1|k-1}
-	\end{equation}
-
-	La covariance de l'erreur de prédiction devient :
-	\begin{align*}
-		P_{k|k-1} &= E\left[\left(A_d\widehat{x}_{k-1|k-1} - x_k\right)\left(A_d\widehat{x}_{k-1|k-1} - x_k\right)^H\right]\\
-		&= E\left\{
-			\left[
-				A_d(\widehat{x}_{k-1|k-1} - x_{k-1}) + (A_dx_{k-1} - x_k)
-			\right]\left[
-			A_d(\widehat{x}_{k-1|k-1} - x_{k-1}) + (A_dx_{k-1} - x_k)
-			\right]^H
-		\right\} \\
-		&= A_dP_{k-1|k-1}Ad^T + Q_d + 0
-	\end{align*}
-
-	Le dernier terme s'explicant par la décorrélation des erreurs sur le modèle d'évolution et d'erreur de prédiction. On obtient finalement le modèle :
-	\begin{equation}
-		P_{k|k-1} = \left(A\ast\varepsilon\right)P_{k-1|k-1}\left(A\ast\varepsilon\right)^T + Q_d
-	\end{equation}
-
-	\subsection{Erreur sur le masque UV}
-
-	On cherche à caractériser l'erreur sur la matrice du plan UV $H$.
-Une manière naïve de faire le calcul serait d'ajouter une erreur gaussienne comme sur la matrice $A$, mais en faisant cela on obtient un modèle très instable et peu réaliste.
-Effectivement, les termes de cette matrices étant :
-	$$
-		H_{j,q} = \exp(-j\frac{2\pi}{\lambda}\Delta z_j\dotproduct\mathbf{I}_q)
-	$$
-	L'erreur sur cette matrice pouvant provenir de différentes sources : mauvaise mesure des distances entre les antennes, mauvais alignement des antennes donc erreur sur les directions, le modèle d'erreur est :
-	$$
-		H_{j,q}^{err} =  \exp(-i\frac{2\pi}{\lambda}(\Delta z_j\dotproduct\mathbf{I}_q + \varepsilon)) =  \exp(-i\frac{2\pi}{\lambda}\Delta z_j\dotproduct\mathbf{I}_q)\exp(-i\frac{2\pi}{\lambda}\varepsilon\dotproduct\mathbf{I}_q)
-	$$
-
-	Soit une rotation du complexe associé a chaque direction. \\
-
-	On implémente donc l'erreur de la manière suivante pour quantifier l'erreur acceptable sur les positions des antennes :
-
-	\begin{minted}{matlab}
-poids = 1e-2;
-z_err = z+ randn(size(z))*poids;
-H_err = matF(J,Pix,z_err,lambda,I);
-	\end{minted}
-
-	On peut ensuite mesurer l'évolution du PSNR dans le temps, pour différents écarts possibles :
-
-%	\begin{figure}[H]
-%		\centering
-%		\includegraphics[width=0.7\linewidth]{Perfs/H5e-3}
-%		\caption{Evolution du PSNR pour $<\varepsilon^2>=5$mm}
-%		\label{fig:h5e-3}
-%	\end{figure}
-%	
-%	\begin{figure}[H]
-%		\centering
-%		\includegraphics[width=0.7\linewidth]{Perfs/He-3_1}
-%		\caption{Evolution du PSNR pour $<\varepsilon^2>=1$mm}
-%		\label{fig:he-3}
-%	\end{figure}
-%	
-%	\begin{figure}[H]
-%		\centering
-%		\includegraphics[width=0.7\linewidth]{Perfs/He-2}
-%		\caption{Evolution du PSNR pour $<\varepsilon^2>=1$cm}
-%		\label{fig:he-2}
-%	\end{figure}
-
-	\begin{figure}[H]
-		\centering
-		\begin{subfigure}{.5\textwidth}
-			\centering
-			\includegraphics[width=0.9\linewidth]{Perfs/He-3_1}
-			\caption{Evolution du PSNR pour $<\varepsilon^2>=1$mm}
-			\label{fig:he-3}
-		\end{subfigure}%
-		\begin{subfigure}{.5\textwidth}
-			\centering
-			\includegraphics[width=0.9\linewidth]{Perfs/H5e-3}
-			\caption{Evolution du PSNR pour $<\varepsilon^2>=5$mm}
-			\label{fig:h5e-3}
-		\end{subfigure}
-		\begin{subfigure}{.5\textwidth}
-			\centering
-			\includegraphics[width=0.9\linewidth]{Perfs/He-2}
-			\caption{Evolution du PSNR pour $<\varepsilon^2>=1$cm}
-			\label{fig:he-2}
-		\end{subfigure}
-		\caption{Mesure du psnr pour différentes erreurs sur la position}
-	\end{figure}
-
-	Les mesures obtenues ont \emph{généralement} cette forme : on obtient parfois des évolutions sans le pic au début, mais la mesure va toujours converger vers une valeur identique tant que $<\varepsilon^2>$ reste constant, en oscillant autour de cette valeur.
-On peut donc tracer l'évolution de l'erreur de reconstruction en fonction de $<\varepsilon^2>$ :
-
-	\begin{figure}[H]
-		\centering
-		\includegraphics[width=0.7\linewidth]{Perfs/perf_errF}
-		\caption{Qualité de reconstruction en fonction de l'erreur}
-	\end{figure}
-
-	\subsubsection{Calculs théoriques}
-
-	On considère la matrice $\widetilde{H}$ construite avec une erreur de mesure sur la position des antennes, c'est à dire 
-	$$
-		\forall j \in \llbracket1;J^2\rrbracket,\:\: \widetilde{\Delta} z_j = \Delta z_j + \boldsymbol{\varepsilon}_j
-	$$
-	On suppose l'erreur $\boldsymbol{\varepsilon}$ suivant une loi normale.\\
-	Les termes de la matrice $\widetilde{H}$ deviennent :
-	$$
-		\widetilde{H}_{j,q} = \exp(-i\frac{2\pi}{\lambda}\widetilde{\Delta}z_j\dotproduct\mathbf{I}_q) = H_{j,q}\times\exp(-i\frac{2\pi}{\lambda}\boldsymbol{\varepsilon}_j\dotproduct\mathbf{I}_q)
-	$$
-
-	On peut donc établir une matrice de correction $\mathcal{C}$ telle que :
-	$$
-		H = \mathcal{C}\odot\widetilde{H}
-	$$
-	\newpage
-
-	On peut vérifier que l'expression de $K$ reste inchangée : on exprime $P_{k|k}$ :
-	\begin{align*}
-		P_{k|k} &= E\left[\left(\widehat{x}_{k|k} - x_k\right)\left(\widehat{x}_{k|k} - x_k\right)^H\right] \\
-		&= E\left[\left(\widehat{x}_{k|k-1} + K(y_k - H\widehat{x}_{k|k-1}) - x_k\right)\left(\widehat{x}_{k|k-1} + K(y_k - H\widehat{x}_{k|k-1}) - x_k\right)^H\right] \\
-		&= E\left[\left(\widehat{x}_{k|k-1} - K(\C\odot\widetilde{H})\widehat{x}_{k|k-1} + Ky_k - x_k\right)\left(\widehat{x}_{k|k-1} - K(\C\odot\widetilde{H})\widehat{x}_{k|k-1} + Ky_k - x_k\right)^H\right] \\
-		&= E\left[\left((\widehat{x}_{k|k-1}-x_k) -K(\C\odot\widetilde{H})(\widehat{x}_{k|k-1}-x_k) + K(y_k - (\C\odot\widetilde{H})x_k)\right)\left((\widehat{x}_{k|k-1}-x_k) -K(\C\odot\widetilde{H})(\widehat{x}_{k|k-1}-x_k) + K(y_k - (\C\odot\widetilde{H})x_k)\right)^H\right]
-	\end{align*}
-
-	On peut faire apparaître les erreurs de mesure $\varepsilon_{mes}$ et de prédiction $\varepsilon_{pred}$ :
-	\begin{align*}
-		\varepsilon_{mes} &= y_k - (\C\odot\widetilde{H})x_k \\
-		\varepsilon_{pred} &= \widehat{x}_{k|k-1} - x_k
-	\end{align*}
-
-	Ces erreurs sont décorrelées, donc les termes correspondant seront nuls :
-	$$
-		E\left[\varepsilon_{mes}\varepsilon_{pred}^H\right] = 0
-	$$
-
-	En rappelant de plus :
-	\begin{align*}
-		P_{k|k-1} &= E\left[\varepsilon_{pred}\varepsilon_{pred}^H\right] \\
-		R &=  E\left[\varepsilon_{mes}\varepsilon_{mes}^H\right]
-	\end{align*}
-
-	Erreurs aux indices k, on peut développer la formule de $P_{k|k}$
-
-	\begin{align*}
-		P_{k|k} &= P_{k|k-1} + K\left(\C\odot \widetilde{H}\right)P_{k|k-1}\left(\Hc\right)^H - P_{k|k-1}\left(K\Hc\right)^H - K\left(\Hc\right)P_{k|k-1} + KRK^H
-	\end{align*}
-
-	Cette forme est bien minimisée par 
-	\begin{equation}
-		K_k^c = P_{k|k-1}\left(\Hc\right)^H\left(\left(\Hc\right)P_{k|k-1}\left(\Hc\right)^H + R\right)^{-1}
-	\end{equation}
-	Forme similaire au gain $K$ calculé sans correction :
-	\begin{equation}
-		K_k = P_{k|k-1}\Ht^H\left(\Ht P_{k|k-1}\Ht^H + R\right)^{-1}
-	\end{equation}
-
-	Par la suite, on omettra les dépendances en $k$ en supposant $P=P_{k|k-1}$ et $K = K_k$.
-
-	On pose l'erreur sur le gain :
-	\begin{equation}
-		\varepsilon_K = K - K^c 
-	\end{equation}
-
-	Ainsi que l'erreur d'estimation induite par l'erreur de masque :
-	\begin{equation}
-		\varepsilon_{\widehat{x}} = \widehat{x}_{k|k} - \widehat{x}_{k|k}^c
-	\end{equation}
-
-	On omettra de la même façon la dépendance en $k$. En développant :
-	\begin{align*}
-		\varepsilon_{\widehat{x}} &= \left[\widehat{x}_{k|k-1} + K(y_k-\Ht\widehat{x}_{k|k-1})\right] - \left[\widehat{x}_{k|k-1} + K^c\left(y_k - \left(\Hc\right)\widehat{x}_{k|k-1}\right)\right] \\
-		&= \left(K-K^c\right)y_k - \left(K\Ht - K^c(\Hc)\right)\widehat{x}_{k|k-1} \\
-		&= \varepsilon_K\left[y_k - \left(\Hc\right)\widehat{x}_{k|k-1}\right] + K\left(\mathbb{1}-\C\right)\odot\Ht\widehat{x}_{k|k-1} \\
-		&= \varepsilon_K(y_k - (\Hc)\left(\varepsilon_{pred} + x_k\right)) + K\left(\mathbb{1}-\C\right)\odot\Ht\widehat{x}_{k|k-1} \\
-		&= \varepsilon_K\left[\left(y_k - (\Hc)x_k\right) - (\Hc)\varepsilon_{pred}\right] + K\left(\mathbb{1}-\C\right)\odot\Ht\widehat{x}_{k|k-1} \\
-		&= \varepsilon_K\left(\varepsilon_{mes} - (\Hc)\varepsilon_{pred}\right) + K\left(\mathbb{1}-\C\right)\odot\Ht\widehat{x}_{k|k-1}
-	\end{align*}
-
-	Où $\mathbb{1}$ désigne l'élément neutre du produit de Hadamard, la matrice dont tous les termes valent 1.\\
-
-	Sous l'hypothèse de l'introduction d'erreur uniquement sur $H$, la matrice $\C\odot\Ht$ est la matrice corrigée et on obtient :
-	$$
-		y_k = (\Hc) x_k \implies \varepsilon_{mes} = 0
-	$$
-	On ne suppose également pas d'erreur sur la matrice d'évolution $A$ :
-	$$
-		\varepsilon_{pred} = 0
-	$$
-
-	D'où :
-	\begin{equation}
-		\varepsilon_{\widehat{x}} = K \left(\mathbb{1} - \C\right)\odot\Ht\widehat{x}_{k|k-1}
-	\end{equation}
-
-	On peut de même évaluer le différentiel sur les covariances :
-	\begin{equation}
-		\Delta P_{k|k} = P_{k|k} - P_{k|k}^c = \left(K^c\left(\Hc\right) - K\Ht\right)P_{k|k-1}
-	\end{equation}
-
-	De plus, si on suppose une erreur de gain sur les antennes, on peut poser les termes :
-	$$
-		\Ht_{j,q} = (1+\varepsilon_{j,q})H_{j,q}
-	$$
-
-	Où $\varepsilon_q$ suit une loi normale, et désigne une erreur sur le gain de l'antenne $j$ dans la direction $q$.\\
-	On peut de même établir une matrice de correction $\B$ telle que :
-	\begin{equation}
-		H = \Hb
-	\end{equation}
-
-	Les erreurs et gain auront alors la même forme
-	\begin{subequations} 
-		\begin{equation}
-			K_k^b = P\left(\Hb\right)^H\left(\left(\Hb\right)P\left(\Hb\right)^H + R\right)^{-1}
-		\end{equation}
-		\begin{equation}
-			\varepsilon_{\widehat{x}} = K\left(\mathbb{1} - \B\right)\odot\Ht\widehat{x}_{k|k-1}
-		\end{equation}
-		\begin{equation}
-			\Delta P_{k|k} = \left(K^b\left(\Hb\right) - K\Ht\right)P_{k|k-1}
-		\end{equation}
-	\end{subequations}
-
-	\subsubsection{Essais expérimentaux}
-
-	On effectue des essais en introduisant une erreur sur la matrice exacte. Plutôt que d'avoir une matrice erronée $\Ht$ connue on aura $H$ exacte connue et :
-	$$
-		\Ht = \mathcal{E}\odot H
-	$$
-	où $\mathcal{E} = \hinv{\C}$ ou $\mathcal{E} = \hinv{\B}$ en fonction de l'erreur que l'on souhaite introduire.
-
-	\paragraph{Erreur de position}
-	On suppose tout d'abord que l'on ne connaît pas exactement la position des antennes du réseau.
-On introduit sur $H$ une matrice de petites rotations définie par la matrice $\mathcal{R} = \hinv{\C}$ plus tôt.
-\\
-	Pour cela, on génère des erreurs directement sur les positions des antennes, puis on peut obtenir $\C$ avec la connaissance de $H$ et $\Ht$ :
-	\begin{equation}
-		\C = H \oslash \Ht
-	\end{equation}
-
-	Cette division est possible car tous les termes sont de norme 1 ou proche de 1 dans le cas de petites variations.
-
-	On effectue 6 réalisations de Kalman sur 8 instants temporels en introduisant une erreur sur la position des antennes.
-On effectue cette opération pour plusieurs erreurs et on peut observer les erreurs $\fnorm{\varepsilon_K}$ et $\fnorm{\varepsilon_{\widehat{x}}}$ :
-
-	\begin{figure}[H]
-		\centering
-		\begin{subfigure}{.5\textwidth}
-			\centering
-			\includesvg[width=.9\linewidth]{Perfs/e_K.svg}
-		\end{subfigure}%
-		\begin{subfigure}{.5\textwidth}
-			\centering
-			\includesvg[width=.9\linewidth]{Perfs/e_x.svg}
-		\end{subfigure}
-		\caption{Erreur induite sur les matrices par un erreur de position}
-	\end{figure}
-
-	On peut également tracer l'erreur commise sur la reconstruction en fonction de l'erreur sur la matrice K :
-
-	\begin{figure}[H]
-		\centering
-		\includesvg[width=0.7\linewidth]{Perfs/x_fK}
-		\caption{Erreur de reconstruction en fonction de l'erreur sur la matrice K}
-	\end{figure}
 
@@
 		\centering
-		\includegraphics[width=.9\linewidth]{image1}
+		\includegraphics[width=.4\linewidth]{image1}
 		\caption{}
@@
 		\centering
-		\includegraphics[width=.9\linewidth]{image2}
+		\includegraphics[width=.4\linewidth]{image2}
 		\caption{}

### 2022-07-11 main_loop.m

@@ -47,10 +47,20 @@ H = matF(J,D,z,lambda,I);
 % X_0 = true_image ;
 % X_0 = MEM(MVDR(A,reshape(nY(:,1),J,J),Mx,My),iMEM);
 % X_0 = normarr(X_0);
-X_0 = vadapted;
+% X_0 = vadapted;
 % tX(:,1) = vadapted;
 
-P_0 = X_0*X_0' - mean(X_0,'all');
+% P_0 = X_0*X_0' - mean(X_0,'all');
+
+% Nouveau calcul X_0 et P_0 exactes
+NX_0 = repmat(vadapted,1,Nreal);
+vX_0 = abs(0.1*randn(D,Nreal) + NX_0);
+errX_0 = vX_0 - NX_0;
+
+X_0 = vX_0(:,randi(Nreal));
+P_0 = (errX_0 * errX_0')/Nreal;
+
+clear NX_0 vX_0 errX_0
 
 % def listes vides
 wK = zeros(D,J^2,N,W);

### 2022-07-11 main_loop.m
ok ig truc exact marchepo

@@ -11,11 +11,10 @@ f = 1e9;
 lambda = c/f;
 dist = 10*lambda/2.1; % J_b antennes espacées de dist 
 N = 7; % itérations Kalman
-W = 1; % nombre de réalisations
+W = 100; % nombre de réalisations
 SNR = 10;
 iMEM = 5; % nombre itérations MEM 
 RATIO = 10e14; % ratio erreur Kalman
-mode = 'CPU';
 Nreal = 1000;
 % A_ev matrice d'évolution définie plus loin (need D ligne 35)
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@ -36,8 +35,8 @@ I = matI(Mx, My);
 A_ev = eye(D);
 A = matA(z,I,D,f,c);
 
-% erreurs = [0.001 0.003 0.005 0.01 0.03 0.05 0.1 0.3];
-erreurs = [0.001 0.3 0.05];
+erreurs = [0.001 0.003 0.005 0.01 0.03 0.05 0.1 0.3];
+% erreurs = [0.001 0.3 0.05];
 npos = numel(erreurs);
 
 % calcul invariants kalman
@@ -47,18 +46,18 @@ H = matF(J,D,z,lambda,I);
 % X_0 = true_image ;
 % X_0 = MEM(MVDR(A,reshape(nY(:,1),J,J),Mx,My),iMEM);
 % X_0 = normarr(X_0);
-% X_0 = vadapted;
+X_0 = vadapted;
 % tX(:,1) = vadapted;
 
-% P_0 = X_0*X_0' - mean(X_0,'all');
+P_0 = X_0*X_0' - mean(X_0,'all');
 
 % Nouveau calcul X_0 et P_0 exactes
-NX_0 = repmat(vadapted,1,Nreal);
-vX_0 = abs(0.1*randn(D,Nreal) + NX_0);
-errX_0 = vX_0 - NX_0;
-
-X_0 = vX_0(:,randi(Nreal));
-P_0 = (errX_0 * errX_0')/Nreal;
+% NX_0 = repmat(vadapted,1,Nreal);
+% vX_0 = abs(0.001*randn(D,Nreal) + NX_0);
+% errX_0 = vX_0 - NX_0;
+% 
+% X_0 = vX_0(:,randi(Nreal));
+% P_0 = (errX_0 * errX_0')/Nreal;
 
 clear NX_0 vX_0 errX_0
 

### 2022-07-24 main_loop.m
Update main_loop.m

@@ -35,7 +35,8 @@ I = matI(Mx, My);
 A_ev = eye(D);
 A = matA(z,I,D,f,c);
 
-erreurs = [0.001 0.003 0.005 0.01 0.03 0.05 0.1 0.3];
+% erreurs = [0.001 0.003 0.005 0.01 0.03 0.05 0.1 0.3];
+erreurs = [0.001 0.003 0.005 0.008 0.01 0.03];
 % erreurs = [0.001 0.3 0.05];
 npos = numel(erreurs);
 
@@ -91,11 +92,11 @@ for err = 1:npos
         ETA = tMoyOp*opRemain;
         fprintf('Erreur : %u/%u -- Réalisation : %u/%u -- Time : %.4f -- ETA : %.4f\n',err,npos,realisation,W,elapsedTime,ETA);
         % 1. Gen données aléatoires
-
-        z_err = z+ randn(size(z))*poids;
-        I_err = I;
-        % I_err(4,:) = I_err(4,:) + randn(size(I_err(4,:)))*poids;
-%         I_err = I + randn(size(I))*poids;
+        z_err = z;
+%         z_err = z+ randn(size(z))*poids;
+%         I_err = I;
+%         I_err(4,:) = I_err(4,:) + randn(size(I_err(4,:)))*poids;
+        I_err = I + randn(size(I))*poids;
         H_err = matF(J,D,z_err,lambda,I_err);
 
 %         C = H ./ H_err;

### 2022-07-24 article/article.tex

@@
 \pagestyle{fancy}
-\lhead{\UE\, - \type\ifthenelse{\nb > 0}{\nb}{}}
+\lhead{Dynamic image reconstruction}
 \rfoot{\thepage /\pageref{LastPage}}
@@
 
-%compteur questions
-\newcounter{question}
-\setcounter{question}{1}
-\newcounter{subq}
-\setcounter{subq}{1}
-
-\newcommand{\rsubq}{\setcounter{subq}{1}}
-\newcommand{\question}{\mgn{\thequestion .}\stepcounter{question}\rsubq}
-\newcommand{\rstq}{\setcounter{question}{1}\rsubq}
-\newcommand{\skipq}[1]{\addtocounter{question}{#1}\rsubq}
-\newcommand{\subq}{\noindent\alph{subq})\stepcounter{subq}} %a modifier pour mettre dans le margin
+\newcommand{\hinv}[1]{#1^{\circ-1}} % inverse de hadamard
+\newcommand{\fnorm}[1]{|\vert#1|\vert_{F}} % norme de Frobenius
 
@@
 
+	\vspace*{.3\textheight}
 	\huge
-	\textbf{Reconstruction dynamique d'images radioastronomiques de trous noirs}
+	\centering
+	\textbf{Dynamic radioastronomic image reconstruction using Kalman filter under ionospheric perturbations}
 
-	\vspace{1.5cm}
+	\vspace{1cm}
 	\LARGE
-	\textbf{Gabriel Robert-Dautun}
+	G. Robert-Dautun
 
@@
 	\large
-	Encadré par Mme Vin
+	Supervised by I. Vin
 
@@
 
-	\includegraphics[width=0.75\textwidth]{logo.PNG}
-
 	\Large
-	Satie
+	Satie\\
 	France\\
@@
 	\newpage
+
+	\part{Introduction}
+
+	\newpage
+	\part{Notations and problem}
+	\section{Notations}
+
+		\subsection{Hadamard product}
+
+	%On introduit la notation $\odot$ associée au produit de Hadamard, qui effectue la multiplication élément à élément de deux matrices de même taille :
+	Hadamard product, denoted $\odot$, is the element-wise product of same sized matrices :
+
+	$$
+	\forall (A,B)\in\left(\mathbb{C}^{m\times n}\right)^2, \quad A\odot B\in\mathbb{C}^{m\times n} \quad\text{and}\quad \forall (i,j)\in\llbracket1;m\rrbracket\times\llbracket1;n\rrbracket,\, (A\odot B)_{i,j} = A_{i,j}\times B_{i,j}
+	$$
+
+	%On introduit également l'inverse de Hadamard, qui inverse élément par élément une matrice
+	We can further introduce Hadamard inversion :
+	$$
+	\forall A \in \left(\mathbb{C}\backslash\{0\}\right)^{m\times n},\quad \hinv{A}\in\left(\mathbb{C}\backslash\{0\}\right)^{m\times n} \quad\text{and}\quad \forall (i,j)\in\llbracket1;m\rrbracket\times\llbracket1;n\rrbracket,\, \left(\hinv{A}\right)_{i,j} = \left(A_{i,j}\right)^{-1}
+	$$
+
+%	Ainsi que la notation $\oslash$ désignant la division de Hadamard :
+	And notation $\oslash$ being Hadamard division :
+	$$
+	\forall (A,B)\in\mathbb{C}^{m\times n}\times\left(\mathbb{C}\backslash0\right)^{m\times n}, \quad A\oslash B\in\mathbb{C}^{m\times n} \quad\text{and}\quad \forall (i,j)\in\llbracket1;m\rrbracket\times\llbracket1;n\rrbracket,\, (A\oslash B)_{i,j} = \frac{A_{i,j}}{B_{i,j}}
+	$$
+
+	\subsection{Norms and distances}
+
+	%On dénote $\fnorm{\bullet}$ la norme de Frobenius :
+	In order to compute distances between matrices, we note $\fnorm{\bullet}$ Frobenius norm :
+	$$
+	A\in\mathcal{M}_{m,n}(\mathbb{K}) \quad \fnorm{A} := \sqrt{\tr{AA^H}} = \sqrt{
+		\;\;\;\sum_{
+			\mathclap{
+				\substack{
+					1\le i\le m \\1\le j\le n
+				}
+			}
+		}\abs{A_{ij}}^2}
+	$$
+
+	%Qui induit sur $\mathcal{M}_{m,n}(\mathbb{D})$ la distance normalisée $d_1$ :
+	Inducing the distance $d_1$ :
+	$$
+	\forall(A,B)\in\left(\mathcal{M}_{m,n}(\mathbb{K})\right)^2\quad d_1(A,B) = \frac{\fnorm{A-B}}{mn}
+	$$
+
+	%Ainsi que la distance $d_2$ :
+	In order to measure influence of a specific element, the ??? distance $d_2$ is the following :
+	$$
+	\forall(A,B)\in\left(\mathcal{M}_{m,n}(\mathbb{K})\right)^2\quad d_2(A,B) = \sum_{
+		\mathclap{
+			\substack{
+				1\le i\le m \\1\le j\le n
+			}
+		}
+	}\abs{A_{ij} - B_{i,j}}
+	$$
+
+	\section{Evolution model and Kalman filtering}
+
+	We consider a set $\{x_k\}$ of $n$ vector as theoretical images taken at a regular time interval.
+Those images are measured by an antenna mesh which UV plane will be written $H$.
+Measurements vectors will be designated by $n$ measures $\{y_k\}$.\\
+	We will consider the following model to describe time related evolutions and measure equation :
+	$$
+		\begin{cases}
+			x_k = Ax_{k-1} + w_{k-1}\\
+			y_k = Hx_k + v_k
+		\end{cases}
+	$$
+
+	Where :
+	\begin{itemize}
+		\item $A$ is the linear evolutionary model between time $k-1$ and $k$, supposing evolution is time invariant
+		\item $\{w_k\}$ are the additive evolutionary errors
+		\item $\{v_k\}$ are the measurements noises
+	\end{itemize}
+
+	In order to build images dynamically with better precision, a Kalman filter has been implemented.\\
+
+	This filter requires assumptions about
+
+	\newpage
+	\part{Reliability of Kalman Filter under perturbations}
+	\section{Perturbations sources}
+	\section{Effects}
+	\subsection{Position}
+	\subsection{Direction}
+
+	\newpage
+	\part{Experimental correction}
+	\section{implementation}
+	\section{Measures}
 

### 2022-07-25 article/article.tex
update

@@
 
+	\subsection{General notations}
+
 		\subsection{Hadamard product}
@@
 	We consider a set $\{x_k\}$ of $n$ vector as theoretical images taken at a regular time interval.
-Those images are measured by an antenna mesh which UV plane will be written $H$.
+Those images are measured by an antenna mesh which UV plane will be written $H$, more generally written as observation model.
 Measurements vectors will be designated by $n$ measures $\{y_k\}$.\\
-	We will consider the following model to describe time related evolutions and measure equation :
+	We will consider the following model to describe time evolution and observation equation :
 	$$
 		\begin{cases}
-			x_k = Ax_{k-1} + w_{k-1}\\
+			x_k = Ax_{k-1} + w_{k}\\
 			y_k = Hx_k + v_k
@@
 	\begin{itemize}
-		\item $A$ is the linear evolutionary model between time $k-1$ and $k$, supposing evolution is time invariant
-		\item $\{w_k\}$ are the additive evolutionary errors
-		\item $\{v_k\}$ are the measurements noises
+		\item $A$ is the linear state-transition model between time $k-1$ and $k$, supposing evolution is time invariant
+		\item $\{w_k\}$ is the state noise sequence
+		\item $\{v_k\}$ is the measurements noise sequence
 	\end{itemize}
@@
 
-	This filter requires assumptions about
+	This filter requires more inputs to estimate the state of the system :
+	\begin{itemize}
+		\item $\{Q_k\}$, the covariance of the process noise sequence
+		\item $\{R_k\}$, the covariance of the observation noise sequence
+	\end{itemize}
+
+	Which can be rewritten as
+	$$
+		\begin{cases}
+			Q_k = \expval{w_{k-1}w_{k-1}^H} - \expval{w_{k-1}}\expval{w_{k-1}}^H \\
+			R_k = \expval{v_{k}v_{k}^H} - \expval{v_k}\expval{v_k}^H \\
+		\end{cases}
+	$$
 

### 2022-07-25 article/article.tex
biblatex config (use biber)

@@
 \usepackage{amssymb,amsmath,minted,float,graphicx,textcomp,systeme,listings,physics,mathtools,ifthen,esvect,hyperref,fancyhdr,lastpage,marginnote,chngcntr,fancyvrb,subcaption,svg}
+
+\usepackage[backend=biber,style=numeric]{biblatex}
+\addbibresource{article.bib}
 
@@
 \newcommand{\fnorm}[1]{|\vert#1|\vert_{F}} % norme de Frobenius
+\newcommand{\moy}[1]{\boldsymbol{\mathsf{m}}_{#1}}
+
+\renewcommand{\expval}[1]{\text{E}\left[#1\right]}
+\newcommand{\w}{\boldsymbol{w}}
+\renewcommand{\v}{\boldsymbol{v}}
+\newcommand{\Q}{\boldsymbol{\mathsf{Q}}}
+\newcommand{\R}{\boldsymbol{\mathsf{R}}}
+
+\newcommand{\x}{\boldsymbol{x}}
+\newcommand{\y}{\boldsymbol{y}}
+
+\newcommand{\xp}{\widehat{\x}_{k|k-1}}
+\newcommand{\xa}{\widehat{\x}_{k-1|k-1}}
+\newcommand{\xe}{\widehat{\x}_{k|k}}
+\newcommand{\yp}{\widehat{\y}_{k|k-1}}
+
+\newcommand{\Pp}{\boldsymbol{\mathsf{P}}_{k|k-1}}
+\newcommand{\Pa}{\boldsymbol{\mathsf{P}}_{k-1|k-1}}
+\newcommand{\Pe}{\boldsymbol{\mathsf{P}}_{k|k}}
 
@@
 
+		A Kalman filter is a linear quadratic estimator, using observations, state-transition model and knowledge about noise and errors to compute a more accurate result than observations alone.
+Radio astronomical measures being prone to noise and errors, usage of this filtering method seems \emph{a priori} a good way to improve accuracy of reconstruction.
+\\
+		One objective of this paper is to evaluate if measure conditions and results are compatible with the filtering.
+
 	\newpage
@@
 	\subsection{General notations}
+
+	$\expval{\bullet}$ denotes the expected value of a random variable/vector/matrix.
+
+	\paragraph{}Bold symbols represent vectors, normal ones represent scalars
 
@@
 		\begin{cases}
-			x_k = Ax_{k-1} + w_{k}\\
-			y_k = Hx_k + v_k
+			\x_k = A\x_{k-1} + \w_{k-1}\\
+			\y_k = H\x_k + \v_k
 		\end{cases}
@@
 	\begin{itemize}
-		\item $A$ is the linear state-transition model between time $k-1$ and $k$, supposing evolution is time invariant
-		\item $\{w_k\}$ is the state noise sequence
-		\item $\{v_k\}$ is the measurements noise sequence
+		\item $A$ is the linear state-transition model between time $k-1$ and $k$, assuming evolution is time invariant
+		\item $\{\w_k\}$ is the state noise sequence
+		\item $\{\v_k\}$ is the measurements noise sequence
 	\end{itemize}
@@
 	\begin{itemize}
-		\item $\{Q_k\}$, the covariance of the process noise sequence
-		\item $\{R_k\}$, the covariance of the observation noise sequence
+		\item $\{\Q_k\}$, the covariance of the process noise sequence
+		\item $\{\R_k\}$, the covariance of the observation noise sequence
 	\end{itemize}
@@
 		\begin{cases}
-			Q_k = \expval{w_{k-1}w_{k-1}^H} - \expval{w_{k-1}}\expval{w_{k-1}}^H \\
-			R_k = \expval{v_{k}v_{k}^H} - \expval{v_k}\expval{v_k}^H \\
+			\Q_k = \expval{\w_{k-1}\w_{k-1}^H} - \expval{\w_{k-1}}\expval{\w_{k-1}}^H \\
+			\R_k = \expval{\v_{k}\v_{k}^H} - \expval{\v_k}\expval{\v_k}^H 
 		\end{cases}
 	$$
+
+	Those inputs are not \emph{a priori} known, but it has been shown that using the \emph{autocovariance least-squares} (ALS) method provides a good estimate of those matrices. [CITATION NEEDED] \\
+	We will assume that these inputs are known perfectly except when specified otherwise.\\
+
+	Kalman filtering works in 2 steps :
+	\paragraph{Prediction} : This step predicts the state of the system at next time-step through de transition-state model, and the predicted measurement :
+	$$
+	\begin{cases}
+		\xp = A\xa + \moy{\w_k}\\
+		\yp = H\xp + \moy{\v_k}
+	\end{cases}
+	$$
+	And the covariance matrix :
+	$$
+		\Pp = H\Pa H^H + \Q_k
+	$$
+	This result can be computed using least-square estimation of the prediction error covariance\cite{intro_KF}
+
+	\paragraph{Update} : Using prior knowledge about noise and observation model, the main goal is to build an image from both prediction and measurement using maximum likelihood estimation to mix them.
 
@@
 
+	\newpage
+	\printbibliography
 \end{document}

### 2022-07-27 article/article.tex
update model + sources

@@
 \newcommand{\moy}[1]{\boldsymbol{\mathsf{m}}_{#1}}
+\newcommand{\autocorr}[1]{\expval{(#1)(#1)^H}}
+\newcommand{\ccorr}[2]{\expval{(#1)(#2)^H}}
 
@@
 \newcommand{\R}{\boldsymbol{\mathsf{R}}}
+\renewcommand{\H}{\boldsymbol{\mathsf{H}}}
+\newcommand{\A}{\boldsymbol{\mathsf{A}}}
 
@@
 \newcommand{\Pe}{\boldsymbol{\mathsf{P}}_{k|k}}
+\renewcommand{\P}{\boldsymbol{\mathsf{P}}}
 
@@
 	$$
-		\Pp = H\Pa H^H + \Q_k
+		\Pp = \autocorr{\xp - \x_k} = H\Pa H^H + \Q_k
 	$$
@@
 
-	\paragraph{Update} : Using prior knowledge about noise and observation model, the main goal is to build an image from both prediction and measurement using maximum likelihood estimation to mix them.
+	\paragraph{Update} : Using prior knowledge about noise and observation model, the main goal is to build an image from both prediction and measurement using maximum likelihood estimation in order to enhance the resulting image.
+This can be written as looking for $K_k$, known as \emph{Kalman gain}, such as :
+	\begin{equation}
+		\xe = \xp + K_k\left(\y_k - \yp\right)
+	\end{equation}
+
+	minimises the quadratic error :
+	\begin{equation}
+		K_k = \arg\min_K \Pe = \arg\min_K \autocorr{\xe - \x_k}
+	\end{equation}
+
+	The resulting expression for $K_k$ is \cite{intro_KF} :
+	\begin{equation}
+		K_k = \Pp H^H\left(H\Pp H^H + \R_k\right)^{-1}
+	\end{equation}
+
+	Thus the update phase can be written :
+	\begin{equation}
+		\begin{cases}
+			\xe = \xp + K_k\left(\y_k - \yp\right) \\
+			\Pe = \left(\mathbb{I} - K_kH\right)\Pp
+		\end{cases}
+	\end{equation}
 
@@
 	\section{Perturbations sources}
+
+	In radio astronomy, many factors can and will affect measures. Those can be :
+	\paragraph{Initialisation error} : Kalman filtering requires initializing the first image and the first correlation matrix. Those cannot be exact and will present errors. \\
+	Concerning the initial image, multiple precise and efficient methods exist \cite{bible}, and from this initial image the initial correlation matrix can ben initialised\cite{intro_KF} with precision as
+	$$
+		\P_0 = \autocorr{\x_0} - \moy{\x_0}\moy{\x_0}^H
+	$$
+
+	\paragraph{Error covariance matrices} : as mentioned above, these matrices set can be estimated with good accuracy using ALS algorithms [CITATION NEEDED].
+It is to be noted that errors on these matrices will not induce new errors in the estimated images, as they are only used to compute the Kalman gain.
+This implies that inaccurate error matrices can only spread already existing errors on estimated images.
+
+	\paragraph{Antennas informations} : Multiple equations in Kalman filtering relies on the observation model $\H$, which is computed as the UV plane FFT covered by the antenna mesh.
+The limited knowledge of this mesh can create errors in $\H$, particularly when antennas positions and directions cannot be perfectly measured.
+
+	\paragraph{Ionospheric perturbations} : Radio astronomical observations can be affected by the passage of the observed electromagnetic waves through the ionosphere \cite{iono}.
+This can affect the observation model $\H$ as it modifies the perceived direction of the electromagnetic wave measured, and calibration or correction must be made to compensate these perturbations.
+
 	\section{Effects}

### 2022-07-28 sim/dz_c.m
lsqnonlin ok c'est lessgo

@@ -1,0 +1,9 @@
+function dz_o = dz_c(J,z)
+dz_o = zeros(J,J,2);
+for j = 1 : J
+    for i = 1:J
+        dz_o(i,j,1:2) = z(j,:)-z(i,:);
+    end
+end
+dz_o = reshape(dz_o,J^2,2);
+end

### 2022-07-28 sim/matF.m

@@ -1,15 +1,10 @@
 function F = matF(J,Q,z,lambda,I)
 %MATF Summary of this function goes here
 %   Detailed explanation goes here
-dz = zeros(J,J,2);
+
 F = zeros(J^2,Q);
 
-for j = 1 : J
-    for i = 1:J
-        dz(i,j,1:2) = z(j,:)-z(i,:);
-    end
-end
-dz = reshape(dz,J^2,2);
+dz = dz_c(J, z);
 
 for q=1:Q
     F(: ,q) = exp(-1i*2*pi/lambda*(dz(:,1)*I(q,1)+dz(:,2)*I(q,2)));

### 2022-07-28 article/article.tex
update article

@@
 
-	\section{Effects}
+	\section{Effects of perturbations}
+
+	These perturbations create errors on the matrices used in reconstruction and prediction with Kalman filtering.
+The goal of this part is to evaluate how much these error affect prediction and reconstruction, to determine if Kalman filtering can be realistically used in radio-astronomic context.
+\\
+
+	\subsection{Methodology}
+
+	In order to measure precisely the effects, the following methodology will be used :
+	\begin{itemize}
+		\item 
+	\end{itemize}
+
 	\subsection{Position}

### 2022-08-01 article/article.tex
art update

@@
 \newcommand{\A}{\boldsymbol{\mathsf{A}}}
+\newcommand{\I}{\boldsymbol{\mathsf{I}}}
 
@@
 \newcommand{\y}{\boldsymbol{y}}
+\newcommand{\z}{\boldsymbol{z}}
+
+\newcommand{\dz}{\Delta\boldsymbol{z}}
 
@@
 \renewcommand{\P}{\boldsymbol{\mathsf{P}}}
+
+\newcommand{\vbeps}{\boldsymbol{\varepsilon}}
+\newcommand{\C}{\mathcal{C}}
 
@@
 
+	When observing distant objects, instruments resolution matters the most.
+When using optical telescopes, the determining factor for resolution is the size of the primary mirror.
+When observing really distant objects, the required size to have a decent resolution becomes unrealistically big.
+To address this problem, radio interferometry replaces a single optical instrument with multiple antennas collecting electromagnetic waves.
+When gathering these measures together, we can obtain a better angular resolution than any real single optical instrument could get.
+
 		A Kalman filter is a linear quadratic estimator, using observations, state-transition model and knowledge about noise and errors to compute a more accurate result than observations alone.
@@
 
-	$\expval{\bullet}$ denotes the expected value of a random variable/vector/matrix.
+	$\expval{\bullet}$ denotes the expected value of a random variable/vector/matrix.\\
 
-	\paragraph{}Bold symbols represent vectors, normal ones represent scalars
+	Bold symbols represent vectors, normal ones represent scalars except when stated otherwise.\\
+
+	Operator $\moy{\bullet}$ represents the mean value/vector/matrix of the given $\bullet$ set.\\
+
+	When a 2-D matrix element is indexed with a single coordinate, it designates the reshaped matrix in a 1-D vector.
 
@@
 
+	\section{Measure and parameters}
+
+	In order to measure precisely the effects, the following parameters will be used :
+	\begin{itemize}
+		\item The antennas distribution is a 8.57m square 6$\times$6 grid. The number of antennas is thus $J = 36$
+		\item The observed directions grid is a 10$\times$10 square. The number of directions is thus $D=100$
+		\item A normally distributed observation noise will be introduced with Signal to Noise Ratio (SNR) of 10. 
+		\item The first image used for Kalman will be computed using Minimum Variance Distortionless Response (MVDR) method \cite{bible}, deconvoluted with Maximum Entropy Method (MEM) \cite{MEM}
+		\item Measured wavelength is $\lambda =0.3$m
+	\end{itemize}
+
+	The following matrices and sets will be used to describe this setup :
+	\begin{itemize}
+		\item $\z$ represents the positions of each antenna in the 2-D grid
+		\item $\dz_{ij}$ is the baseline : $\dz_{ij} = \z_i - \z_j$. 
+		\item $\I$ is the directions matrix, represented as normalised direction cosines.
+	\end{itemize}
+
 	\section{Evolution model and Kalman filtering}
 
-	We consider a set $\{x_k\}$ of $n$ vector as theoretical images taken at a regular time interval.
-Those images are measured by an antenna mesh which UV plane will be written $H$, more generally written as observation model.
-Measurements vectors will be designated by $n$ measures $\{y_k\}$.\\
+	We consider a set $\{\x_k\}$ of $n$ vector as theoretical images taken at a regular time interval.
+Those images are measured by an antenna mesh $\{\z_k\}$ which UV plane will be written $H$, more generally written as observation model.
+Measurements vectors will be designated by $n$ measures $\{\y_k\}$.\\
+	$\H$ can be computed by the following :
+	\begin{equation}
+		\H_{j,q} = \exp(-i\frac{2\pi}{\lambda}\dz_{j}\dotproduct\I_q)
+	\end{equation}
 	We will consider the following model to describe time evolution and observation equation :
@@
 		\begin{cases}
-			\x_k = A\x_{k-1} + \w_{k-1}\\
-			\y_k = H\x_k + \v_k
+			\x_k = \A\x_{k-1} + \w_{k-1}\\
+			\y_k = \H\x_k + \v_k
 		\end{cases}
@@
 	Concerning the initial image, multiple precise and efficient methods exist \cite{bible}, and from this initial image the initial correlation matrix can ben initialised\cite{intro_KF} with precision as
-	$$
+	\begin{equation}\label{p0}
 		\P_0 = \autocorr{\x_0} - \moy{\x_0}\moy{\x_0}^H
-	$$
+	\end{equation}
 
@@
 
-	\subsection{Methodology}
+	\subsection{Methodology and Parameters}
 
-	In order to measure precisely the effects, the following methodology will be used :
+	In order to simulate data, an image is introduced as the "reference image", this image will only be used in order to compute simulated observations.
+The selected picture must respect the following conditions :
 	\begin{itemize}
-		\item 
+		\item mostly black background, to give realistic results for space imagery
+		\item no convolution effect, as all directions are considered sources of different intensities, only real sources must appear
 	\end{itemize}
+	The following picture is the one used :
+
+	\begin{figure}[H]
+		\centering
+		\includegraphics[width=0.5\linewidth]{src/bh_square}
+		\caption{Selected image for simulation}
+		\label{fig:bhsquare}
+	\end{figure}
+
+	Different images gave similar results if they followed the conditions cited above.\\
+
+	First the basic parameters and matrices are computed : antenna grid, directions cosines, baseline vectors.\\ Next, perfect base matrices are computed : errorless observation matrix $\H$, reference observations using the errorless $\H$, error covariance matrices.
+The latter are computed perfectly using known simulated noise, and errors from ALS estimation are neglected.
+A set of perfect images are also computed to measure reconstruction absolute error.\\
+	Initial image $\x_0$ and covariance $\P_0$ are computed as stated in equation (\ref{p0}).
 
 	\subsection{Position}
+
+	The first error studied is the position error. A normally distributed error $\vbeps$ is introduced in the antennas positions matrix, giving :
+
+	$$
+		\z_k^* = \z_k + \vbeps_k
+	$$
+
+	The new inaccurate baseline is :
+	$$
+		\dz_{ij}^* = \z_i + \vbeps_i - \z_j - \vbeps_j = \dz_{ij} + \vbeps_{ij}
+	$$
+
+	The new inaccurate observations matrix's expression is the following :
+	\begin{equation}
+		\H_{j,q}^* = \exp(-i\frac{2\pi}{\lambda}\dz_{j}^*\dotproduct\I_q) = \exp(-i\frac{2\pi}{\lambda}\left(\dz_{j} + \vbeps_j\right)\dotproduct\I_q) = \exp(-i\frac{2\pi}{\lambda}\dz_{j}\dotproduct\I_q)\exp(-i\frac{2\pi}{\lambda}\vbeps_j\dotproduct\I_q)
+	\end{equation}
+
+	The distortion matrix $\C$ can thus be defined such as :
+	\begin{equation}
+		\H^* = \H\odot\C 
+	\end{equation}
+
 	\subsection{Direction}

### 2022-08-01 article/article.tex
art update

@@
 \newcommand{\I}{\boldsymbol{\mathsf{I}}}
+\newcommand{\K}{\boldsymbol{\mathsf{K}}}
 
@@
 
-	When a 2-D matrix element is indexed with a single coordinate, it designates the reshaped matrix in a 1-D vector.
+	When a 2-D matrix element is indexed with a single coordinate, it designates the reshaped matrix in a 1-D vector.\\
+
+	$\mathbb{I}$ is the identity matrix.
 
@@
 	$$
-		\Pp = \autocorr{\xp - \x_k} = H\Pa H^H + \Q_k
+		\Pp = \autocorr{\xp - \x_k} = \A\Pa \A^H + \Q_k
 	$$
@@
 
+	By the same calculation than before \cite{intro_KF}, position error being decorrelated with other errors or noise, the equations become :
+	\begin{subequations}
+		\begin{equation}\label{xpe}
+			\xp^* = \A\xa + \moy{\w_k}
+		\end{equation}\begin{equation}
+			\yp^* = \H^*\xp^* + \moy{\v_k}
+		\end{equation}\begin{equation}
+			\Pp^* = \A\Pa^*\A^H + \Q_k
+		\end{equation}\begin{equation}
+			\K_k^* = \Pp^*\left(\H^*\right)^H\left(\H^*\Pp^*\left(\H^*\right)^H\right)^{-1}
+		\end{equation}\begin{equation}
+			\Pe^* = \Pp^* + \K_k^*\H^*\Pp^*\left(\H^*\right)^H\left(\K_k^*\right)^H + \K_k^*\R_k\left(\K_k^*\right)^H
+		\end{equation}\begin{equation}
+			\xe^* = \xp^* + \K_k^*(\y_k - \yp^*)
+		\end{equation}
+	\end{subequations}
+
+	To evaluate reconstruction quality, we can introduce the following errors :
+	\begin{equation}
+		\vbeps^{\K}_k = \K_k^* - \K_k
+	\end{equation}
+	\begin{equation}\label{errX}
+		\vbeps^{\x}_k = \xe^* - \xe
+	\end{equation}
+
+	Respectively gain error and estimation error.\\
+	Using equation (\ref{xpe}), (\ref{errX}) can be rewritten :
+	\begin{equation}
+		\vbeps^{\x}_k = \vbeps^{\K}_k\y_k + \left(\mathbb{I} - \K_k^*\H^*\right)\xp^* - \left(\mathbb{I} - \K_k\H\right)\xp
+	\end{equation}
+
+	To evaluate the reconstruction error evolution
+
 	\subsection{Direction}

### 2022-08-02 article/article.tex
one direction >>>

@@
 \newcommand{\C}{\mathcal{C}}
+\newcommand{\D}{\mathcal{D}}
 
@@
 When gathering these measures together, we can obtain a better angular resolution than any real single optical instrument could get.
+\\
+
+	Those distant objects are not fixed in time and a lot of calculations have been done to create evolution models for them.
+The objective here is not to use observations to confirm models, but to assume a model is known and get better results from the knowledge of observations and model.
 
@@
 A set of perfect images are also computed to measure reconstruction absolute error.\\
-	Initial image $\x_0$ and covariance $\P_0$ are computed as stated in equation (\ref{p0}).
+	Initial image $\x_0$ and covariance $\P_0$ are computed as stated in equation (\ref{p0}).\\
+
+	The next step is the generation of an inaccurate matrix normally distributed around the true one - position or direction here.
+From this matrix is generated an inaccurate  $\H^*$ observations matrix if needed, and Kalman filter is run two times, the first using perfect parameters and the second one using inaccurate ones.
+In order to smooth out random noise effects this step is repeated 100 times.
+\\
+
+	We can then get the mean error values, measured using a distance defined above, and standard deviations needed.
 
@@
 
-	The first error studied is the position error. A normally distributed error $\vbeps$ is introduced in the antennas positions matrix, giving :
+	The first error studied is the position error. A normally distributed error $\vbeps$ is introduced on the antennas positions matrix, giving :
 
@@
 			\Pe^* = \Pp^* + \K_k^*\H^*\Pp^*\left(\H^*\right)^H\left(\K_k^*\right)^H + \K_k^*\R_k\left(\K_k^*\right)^H
-		\end{equation}\begin{equation}
+		\end{equation}\begin{equation}\label{finKal}
 			\xe^* = \xp^* + \K_k^*(\y_k - \yp^*)
@@
 
-	To evaluate the reconstruction error evolution
+	Computing these equations gives the following results :
+
+	% figure
+
+	% discussion
 
 	\subsection{Direction}
+
+	Direction error can be caused by a misaligned antenna, or a small displacement, but will mostly come from ionospheric perturbation\cite{iono}.
+To study the effects of these perturbations, a normally distributed error $\vbeps$ is introduced on the direction matrix :
+
+	\begin{equation}
+		\I_q^* = \I_q + \vbeps_q
+	\end{equation}
+
+	The observations matrix becomes :
+	$$
+		\H_{j,q}^* = \exp(-i\frac{2\pi}{\lambda}\dz_{j}\dotproduct\I_q^*) = \exp(-i\frac{2\pi}{\lambda}\dz_{j}\dotproduct\I_q)\exp(-i\frac{2\pi}{\lambda}\dz_{j}\dotproduct\vbeps_q)
+	$$
+	A directional distortion matrix can be defined such as :
+	\begin{equation}
+		\H^* = \H \odot \D
+	\end{equation}
+
+	Kalman equations from (\ref{xpe}) to (\ref{finKal}) stay the same.\\
+
+	To further measure the effect of this error, it will be introduced on a single direction first, then on 5 directions, then on all of them.
+
+	\subsubsection{One direction}
 
@@
 	\section{Measures}
+
+	\newpage
+	\part{Further insights}
+
+	% err dir 1 antenne plutôt que sur 1 dir
+	% smoother
+	% deep
 

### 2022-08-07 article/article.tex
update

@@
 	\newpage
+	\part{SATIE}
+
+	\newpage
 	\part{Notations and problem}
@@
 	%Ainsi que la distance $d_2$ :
-	In order to measure influence of a specific element, the ??? distance $d_2$ is the following :
+	In order to measure influence of a specific element, the additive distance $d_2$ is the following :
 	$$
@@
 
+		\section{Data simulation and Image reconstruction}
+	\subsection{Simulation}
+
+	In order to simulate data, an image is introduced as the "reference image", this image will only be used in order to compute simulated observations.
+The selected picture must respect the following conditions :
+	\begin{itemize}
+		\item mostly black background, to give realistic results for space imagery
+		\item no convolution effect, as all directions are considered sources of different intensities, only real sources must appear
+	\end{itemize}
+	The following picture is the one used :
+
+	\begin{figure}[H]
+		\centering
+		\includegraphics[width=0.5\linewidth]{src/bh_square}
+		\caption{Selected image for simulation}
+		\label{fig:bhsquare}
+	\end{figure}
+
+	Different images gave similar results if they followed the conditions cited above.\\
+
+	Data simulation will be done as stated in [REF TER].
+First the visibilities matrix is computed using the reference image and the UV-plane of the antenna mesh.
+To simulate visibilities at different snapshots, we apply the wanted transformation without noise and use the same UV-plane matrix to compute visibilities.\\
+
+	To add noise, a normally distributed random matrix is generated and the covariance of that matrix is added to the visibilities one.
+This matrix could also be directly generated using Wishart distribution.
+\\
+
+	\subsection{Reconstruction}
+
+	To obtain the initial image $\x_0$ two methods are used depending on the goal of the measure. \\
+
+	When reviewing performances under certain errors, all other inputs of the filter - including $\x_0$ - must be errorless.
+The simplest way of doing that for $\x_0$ is to use the reference image as input.
+\\
+
+	When reviewing Kalman filter performances under real conditions, the reference image is not considered as knwon, and we need to reconstruct the image from the visibilities matrix.
+While different methods exist [REF TER], the ones used here are MVDR reconstructions followed by Maximum Entropy deconvolution.
+
 	\newpage
@@
 	\subsection{Methodology and Parameters}
-
-	In order to simulate data, an image is introduced as the "reference image", this image will only be used in order to compute simulated observations.
-The selected picture must respect the following conditions :
-	\begin{itemize}
-		\item mostly black background, to give realistic results for space imagery
-		\item no convolution effect, as all directions are considered sources of different intensities, only real sources must appear
-	\end{itemize}
-	The following picture is the one used :
-
-	\begin{figure}[H]
-		\centering
-		\includegraphics[width=0.5\linewidth]{src/bh_square}
-		\caption{Selected image for simulation}
-		\label{fig:bhsquare}
-	\end{figure}
-
-	Different images gave similar results if they followed the conditions cited above.\\
 
@@
 
-	To further measure the effect of this error, it will be introduced on a single direction first, then on 5 directions, then on all of them.
+	To further measure the effect of this error, it will be introduced on a single direction first, then on all of them.
 
 	\subsubsection{One direction}
+
+	\subsubsection{All directions}
 

### 2022-08-07 main_loop.m

@@ -39,7 +39,7 @@ A = matA(z,I,D,f,c);
 % erreurs = [0.001 0.003 0.005 0.01 0.03 0.05 0.1 0.3];
 % erreurs = [0.001 0.003 0.005 0.008 0.01 0.03];
 % erreurs = [0.001 0.3 0.05];
-erreurs = [0.01];
+erreurs = [0.001];
 npos = numel(erreurs);
 
 % calcul invariants kalman
@@ -47,7 +47,8 @@ H = matF(J,D,z,lambda,I);
 
 [nY,nR,nQ,tX] = dataGen_im(H,J,N,D,vadapted,A_ev,Nreal,SNR,RATIO);
 % X_0 = true_image ;
-X_mvdr = MEM(MVDR(A,reshape(nY(:,1),J,J),Mx,My),iMEM);
+% X_mvdr = MEM(MVDR(A,reshape(nY(:,1),J,J),Mx,My),iMEM);
+X_mvdr = MVDR(A,reshape(nY(:,1),J,J),Mx,My);
 X_mvdr = normarr(abs(X_mvdr));
 X_0 = vadapted;
 % tX(:,1) = vadapted;
@@ -104,26 +105,34 @@ for err = 1:npos
         % Calcul Hcorr
         % 1. poser pb
         
-        y = H_err*X_0;
-        Iv = I(:);
+%         y = H_err*X_mvdr;
+        y = nY(:,1);
+        
+        Iv = reshape(I_err,[],1);
 
         prb = @(x)prb_f(x,y,Iv,dz,J,D,lambda);
 
 
         % 2. init
-        x0 = [zeros(1,2*D) reshape(X_mvdr,1,[])];
+        dbiasPrio = zeros(1,2*D)*err/100;
+        xparfait = I_err-I;
+        dbiasPrioInit = reshape(xparfait,1,2*D) + dbiasPrio;
+        x0 = [dbiasPrioInit reshape(X_mvdr,1,[])];
         lb = [-0.1*ones(1,2*D) zeros(1,D)];
         ub = [0.1*ones(1,2*D) ones(1,D)];
 
         % 3. lsqnonlin
-
-        x = lsqnonlin(prb,x0,lb,ub);
+        options = optimoptions(@lsqnonlin,'Display','Iter','FiniteDifferenceType','central');
[…]

### 2022-08-07 utils/amel.m

@@ -1,0 +1,6 @@
+function res = amel(avant,after,ref)
+%AMEL Summary of this function goes here
+%   Detailed explanation goes here
+res = fro(avant-ref) - fro(after-ref);
+end
+

### 2022-08-07 article/article.tex
update article

@@
 	Those distant objects are not fixed in time and a lot of calculations have been done to create evolution models for them.
-The objective here is not to use observations to confirm models, but to assume a model is known and get better results from the knowledge of observations and model.
+The goal here is not to use observations to confirm models, but to assume a model is known and get better results considering the knowledge of both observations and model.
+\\
 
-		A Kalman filter is a linear quadratic estimator, using observations, state-transition model and knowledge about noise and errors to compute a more accurate result than observations alone.
+	A (non extended) Kalman filter is a linear quadratic estimator, using observations, state-transition model and knowledge about noise and errors to compute a more accurate result than observations alone.
 Radio astronomical measures being prone to noise and errors, usage of this filtering method seems \emph{a priori} a good way to improve accuracy of reconstruction.
 \\
-		One objective of this paper is to evaluate if measure conditions and results are compatible with the filtering.
+
+	A sensitive part of the filter is the observations model, needing precise knowledge of antennas positions and directions.
+An objective of this work is to determine how inaccuracies in these measurements affect image reconstruction, and if it is realistically usable in radio astronomical imagery.
+\\
 
@@
 	\part{SATIE}
+
+	SATIE stands for \emph{laboratoire des Systèmes et Applications des Technologies de l'Information et de l'Energie} (Information and Energy Technology Systems and Applications Laboratory).\\
+	Created in 1975 under the name \emph{Laboratoire d'Electricité de l'ENSET}, then renamed \emph{LESiR} in 1984 at the time of its association with CNRS, then finally SATIE in 2002, the laboratory studies
 
@@
 
-	Data simulation will be done as stated in [REF TER].
+	Data simulation will be done as stated in \cite{TER}.
 First the visibilities matrix is computed using the reference image and the UV-plane of the antenna mesh.
@@
 	When reviewing Kalman filter performances under real conditions, the reference image is not considered as knwon, and we need to reconstruct the image from the visibilities matrix.
-While different methods exist [REF TER], the ones used here are MVDR reconstructions followed by Maximum Entropy deconvolution.
+While different methods exist \cite{TER}, the ones used here are MVDR reconstructions followed by Maximum Entropy deconvolution.
 
@@
 
+	\paragraph{Directional error on one antenna :} on larger arrays, ionospheric perturbations might not affect all antennas the same, thus studying the reliability when one antenna has errors might be an idea to explore.
+
+	\paragraph{Use of smoothers :} when using a Kalman filter, only prior knowledge of data is needed.
+When all measures at different snapshots are done, a \emph{smoother}\cite{RTS} can be used to give better Kalman results, their reliability under perturbations needs to be discussed.
+
+	\paragraph{Deep learning :} In image processing, deep learning and artificial intelligence are getting more effective and reliable.
+In particular, model based deep learning \cite{deep} can be an effective way of rebuilding images when an evolution model is known.
+
 	\newpage

### 2022-08-07 article/article.tex
critic update

@@
 	SATIE stands for \emph{laboratoire des Systèmes et Applications des Technologies de l'Information et de l'Energie} (Information and Energy Technology Systems and Applications Laboratory).\\
-	Created in 1975 under the name \emph{Laboratoire d'Electricité de l'ENSET}, then renamed \emph{LESiR} in 1984 at the time of its association with CNRS, then finally SATIE in 2002, the laboratory studies
+	Created in 1975 under the name \emph{Laboratoire d'Electricité de l'ENSET}, then renamed \emph{LESiR} in 1984 at the time of its association with CNRS, then finally SATIE in 2002, the laboratory covers a large field of study including signal processing, electronics, and energy conversion.\\
+
+	The SATIE is linked to the following academic structures : ENS Paris-Saclay, ENS Rennes, Université Paris-Saclay, Cergy Paris Université, Université Gustave Eiffel, CNAM Paris and CNRS.
+It counts five geographical sites near the aforementioned structures.
+
+	\begin{figure}[H]
+		\centering
+		\includegraphics[width=0.7\linewidth]{src/sites_geoo_satie}
+		\caption{Geographic sites of SATIE (copyright E. Vourc'h)}
+		\label{fig:sitesgeoosatie}
+	\end{figure}
 

### 2022-08-14 article/article.tex
forgot ctrl s lmao
._.

@@
 		\newpage
+	\part{SATIE}
 
+	SATIE stands for \emph{laboratoire des Systèmes et Applications des Technologies de l'Information et de l'Energie} (Information and Energy Technology Systems and Applications Laboratory).\\
+	Created in 1975 under the name \emph{Laboratoire d'Electricité de l'ENSET}, then renamed \emph{LESiR} in 1984 at the time of its association with CNRS, then finally SATIE in 2002, the laboratory covers a large field of study including signal processing, electronics, and energy conversion.\\
+
+	The SATIE is linked to the following academic structures : ENS Paris-Saclay, ENS Rennes, Université Paris-Saclay, Cergy Paris Université, Université Gustave Eiffel, CNAM Paris and CNRS.
+It counts five geographical sites near the aforementioned structures.
+
+	\begin{figure}[H]
+		\centering
+		\includegraphics[width=0.7\linewidth]{src/sites_geoo_satie}
+		\caption{Geographic sites of SATIE (copyright E. Vourc'h)}
+		\label{fig:sitesgeoosatie}
+	\end{figure}
+
+	Research in SATIE is dispatched around two main divisions :
+
+	\begin{tabular}{|c||c|c|} % a traduire ffs
+		\hline
+		Divisions & Components and Systems for electrical energy & Multi-scale information and analysis systems \\ & (CSEE) & (SIAME) \\\hline
+		Research fields & \begin{tabular}{@{}c@{}}
+			Materials: from synthesis to application\vspace{8px} \\ Power electronics: integration, \\ usage constraints, interfaces\vspace{8px} \\ Transducers and energy systems: \\ methodologies, design and control
+		\end{tabular} & \begin{tabular}{@{}c@{}}
+			Instrumentation and imagery\vspace{5px} \\ Methods and tools for signals and systems\\ (MOSS)
+		\end{tabular}
+		\\\hline
+	\end{tabular}
+	\vspace{7px}\\
+	I did my internship in the SIAME/MOSS division. \\
+	The SIAME/MOSS division is focused on getting a better understanding and mastering complex systems. The three main topics of the division are :
+	\paragraph{Data and image analysis:} this topic's main goal is to estimate physical or semantic quantities from observations data.
+This data is incomplete, inaccurate, and not absolutely reliable.
+When data is generated using multi-sensor systems, it presents similarities and complementarities that can be used to get better estimations.
+
+	\newpage
 	\part{Introduction}
@@
 	A sensitive part of the filter is the observations model, needing precise knowledge of antennas positions and directions.
-An objective of this work is to determine how inaccuracies in these measurements affect image reconstruction, and if it is realistically usable in radio astronomical imagery.
+A of this goal work is to determine how inaccuracies in these measurements affect image reconstruction, and if it is realistically usable in radio astronomical imagery.
 \\
-
-	\newpage
-	\part{SATIE}
-
-	SATIE stands for \emph{laboratoire des Systèmes et Applications des Technologies de l'Information et de l'Energie} (Information and Energy Technology Systems and Applications Laboratory).\\
-	Created in 1975 under the name \emph{Laboratoire d'Electricité de l'ENSET}, then renamed \emph{LESiR} in 1984 at the time of its association with CNRS, then finally SATIE in 2002, the laboratory covers a large field of study including signal processing, electronics, and energy conversion.\\
-
-	The SATIE is linked to the following academic structures : ENS Paris-Saclay, ENS Rennes, Université Paris-Saclay, Cergy Paris Université, Université Gustave Eiffel, CNAM Paris and CNRS.
-It counts five geographical sites near the aforementioned structures.
-
-	\begin{figure}[H]
-		\centering
-		\includegraphics[width=0.7\linewidth]{src/sites_geoo_satie}
-		\caption{Geographic sites of SATIE (copyright E. Vourc'h)}
-		\label{fig:sitesgeoosatie}
-	\end{figure}
 

### 2022-08-15 article/article.tex
add graphs + start desc

@@
 
-	\subsection{Methodology and Parameters}
+	\subsection{Methodology details}
 
@@
 
+	In order to measure the performances under positional inaccuracies, the distance $d_1$ is used to evaluate the error between matrices and images :
+	\begin{itemize}
+		\item Reconstruction error : 
+		$$
+			\varepsilon_k^{\x} = d_1\left(\xe^*,\x_k\right)
+		$$
+		\item Additional error : 
+		$$
+			\varepsilon_k^{\x\x} = d_1\left(\xe^*,\xe\right)
+		$$
+	\end{itemize}
+
+	Even if $\{\x_k\}$ are real numbers, computing $\{\xe\}$ using $\H$ yields complex matrices.
+This induces that these errors are not ordered : we can get results like $\varepsilon_k^{\x} < \varepsilon_k^{\x\x}$.
 	Computing these equations gives the following results :
 
-	% figure
+	% figures
+
+	\begin{figure}[H]
+		\centering
+		\begin{subfigure}{.5\textwidth}
+			\centering
+			\includesvg[width=.9\linewidth]{graphs/positions_x_e.svg}
+			\caption{Distance to true image}
+			\label{fig:p_e_x_e}
+		\end{subfigure}%
+		\begin{subfigure}{.5\textwidth}
+			\centering
+			\includesvg[width=.9\linewidth]{graphs/positions_xx.svg}
+			\caption{Distance to "perfect" reconstruction}
+		\end{subfigure}
+	\caption{Simulation results for positional inaccuracy}
+	\end{figure}
 
 	% discussion
+	For these simulations, reconstruction error without inaccuracies is 
+	$$ % expval ? => ptet mettre <.> ?
+	\left\langle d_1\left(\xe,\x_k\right)\right\rangle = 8,46\times10^{-4}
+	$$
+	and is represented by the blue line in (\ref{fig:p_e_x_e}).\\
+
+	The first interesting observations on these graphs are the extremities convergences : above 0,05m position error standard deviation, errors approach a distance of 0,1.
+The $d_1$ distance being normalised (as $\x$ and $\xe$ are normalised between 0 and 1), this distance means 10\% reconstruction error.
+This error level represents non reliable reconstruction.
+It can be illustrated by the following reconstruction :
+	\begin{figure}[H]
+		\centering
+		\begin{subfigure}{.5\textwidth}
+			\centering
+			\includegraphics[width=.7\linewidth]{src/t_im.png}
+			\caption{True image}
+		\end{subfigure}%
+		\begin{subfigure}{.5\textwidth}
+			\centering
+			\includegraphics[width=.7\linewidth]{src/t_err.png}
+			\caption{Reconstruction at 0,1 distance of the true image}
+		\end{subfigure}
+		\caption{Image reconstruction at unreliable distance}
+	\end{figure}
+
+	This distance around 0,1 is actually expected, as the mean distance $d_1$ between two random matrices in $\mathcal{M}_{10}([0;1])$ can be computed and gives 1,167.
+The random effects of position errors getting more important, the reconstruction using this inaccurate positions matrix is becoming random thus converging to this value of 1,167.
 

### 2022-08-15 article/article.tex
dir4 graphs

@@
 			\includegraphics[width=.7\linewidth]{src/t_im.png}
-			\caption{True image}
+			\caption{Errorless reconstruction}
 		\end{subfigure}%
@@
 
-	This distance around 0,1 is actually expected, as the mean distance $d_1$ between two random matrices in $\mathcal{M}_{10}([0;1])$ can be computed and gives 1,167.
-The random effects of position errors getting more important, the reconstruction using this inaccurate positions matrix is becoming random thus converging to this value of 1,167.
+	This distance around 0,1 is actually expected, as the mean distance $d_1$ between two random matrices in $\mathcal{M}_{10}([0;1])$ can be computed and gives 1,167.\footnote{computed as the mean $d_1$ distance between two randomly generated matrices using matlab function \mintinline{matlab}|rand|, repeated $10^4$ times} The random effects of position errors getting more important, the reconstruction using this inaccurate positions matrix is becoming random thus converging to this value of 1,167.
+\\
+
+	On the other side of the graph, the distance seems to converge to a value around $3\times10^{-4}$.
+This value is rather unexpected, as it is below the errorless reconstruction error.
+This may be explained by the complex nature of the reconstructed image, but no certain explanation was found.
+The convergence part is also uncertain, as the mean values seems to converge but the standard deviation expands.
+The convergence at non-zero distance of the errorless reconstruction is also unexplained.
 
@@
 
+	Ionospheric perturbations being local and direction-dependent, it can be interesting to measure the effect of a single direction being inaccurate.
+To evaluate the reconstruction error, the $d_2$ distance is used, as it is additive: the total distance between two reconstructions is the sum of the distance of each direction.
+
 	\subsubsection{All directions}
@@
 	\section{Measures}
+
+	\newpage
+	\part{Conclusion}
 

### 2022-08-24 article/article.tex
Update article.tex
almost there

@@
 \newcommand{\z}{\boldsymbol{z}}
+\renewcommand{\u}{\boldsymbol{u}}
 
@@
 		\includegraphics[width=0.7\linewidth]{src/sites_geoo_satie}
-		\caption{Geographic sites of SATIE (copyright E. Vourc'h)}
+		\caption{Geographic sites of SATIE (\textcopyright E. Vourc'h)}
 		\label{fig:sitesgeoosatie}
@@
 When data is generated using multi-sensor systems, it presents similarities and complementarities that can be used to get better estimations.
+\\
+	The applications are :
+	\begin{itemize}
+		\item Geolocation
+		\item Non-destructive testing
+		\item radio telescopes calibration
+		\item image analysis for shape detection
+	\end{itemize}
+
+	\paragraph{Algorithms-architectures-uses adequacy:} This topic's goal is the design of autonomous systems with matching algorithms and algorithms.
+They develop tools and methods for multi-sensor instrumentation, embedded computing architectures and data flux synchronisation.
+Their work is divided in 4 points :
+
+	\paragraph{Perception systems for mobility:} The last research axis is mostly aimed at the development of autonomous vehicles.
+Its goal is to conceive the hardware part of a perception system, and the software analysis sequence to optimise the system.
+\\
+
+	The SATIE places a great importance on experimentation, and notably concerning the autonomous vehicles conception, as all results must be confirmed with experimentation, the lab has many electrical cars mounted with instruments and automated controls.
+\\
+
+	This organisation can be summarised as follows :
+
+	\begin{figure}[H]
+		\centering
+		\includegraphics[width=.7\linewidth]{src/satie_org}
+		\caption{lab's organisation}
+	\end{figure}
+
+	The lab is also involved in many masters in Université Paris-Saclay, ENS Paris-Saclay, or ENS Rennes etc.
+These masters cover the topics: signal processing; electronics, sensors; electrical energy; mechatronics and higher education (M2FESUP).
+It also accepts some PhD students and interns.
 
@@
 
+	\begin{figure}[H]
+		\centering
+		\includegraphics[width=.5\linewidth]{src/radioastro}
+		\caption{Radio telescope in New Mexico, Very Large Array (VLG)}
+	\end{figure}
+
 	Those distant objects are not fixed in time and a lot of calculations have been done to create evolution models for them.
@@
 
-	A (non extended) Kalman filter is a linear quadratic estimator, using observations, state-transition model and knowledge about noise and errors to compute a more accurate result than observations alone.
+	A Kalman filter is a linear quadratic estimator, using observations, state-transition model and knowledge about noise and errors to compute a more accurate result than observations alone.
 Radio astronomical measures being prone to noise and errors, usage of this filtering method seems \emph{a priori} a good way to improve accuracy of reconstruction.
@@
 \\
+
+	As of today, many optimisations have been done to improve Kalman filter, such as smoothers \cite{RTS}, and models to describe accretion disks evolution.
+This raises the question about the feasibility of using a Kalman filter to enhance radio astronomic pictures of dynamic systems.
 
@@
 
+	Similar to $d_1$, but better known, to measure the quality of image reconstruction, the Peak to Signal Noise Ratio (PSNR) can be computed :
+
+	$$
+		PSNR = 10\cdot\log_{10}\left(\frac{d^2}{EQM}\right)
+	$$
+
+	where $EQM$ is the mean quadratic error between reference image $I_{ref}$ and reconstruction $I_{rec}$ :
+
+	$$
+		EQM = \frac{1}{M_xM_y}\sum_{i=0}^{M_x-1}\sum_{j=0}^{M_y-1}( I_{ref}(i,j) - I_{rec}(i,j) )^2
+	$$
+
+	$d$ is the dynamic of each pixel (here 1), $M_X$ and $M_Y$ the size of the image along respectively axis $X$ and $Y$ in pixels.
+
 	\section{Measure and parameters}
@@
 	\end{itemize}
+
+	\begin{figure}[H]
+		\centering
+		\includegraphics[width=.5\linewidth]{src/antennes}
+		\caption{Illustration of the antennas mesh}
+	\end{figure}
 
@@
 
-	Those inputs are not \emph{a priori} known, but it has been shown that using the \emph{autocovariance least-squares} (ALS) method provides a good estimate of those matrices. [CITATION NEEDED] \\
+	Those inputs are not \emph{a priori} known, but it has been shown that using the \emph{autocovariance least-squares} (ALS) method provides a good estimate of those matrices. \cite{ALS} \\
 	We will assume that these inputs are known perfectly except when specified otherwise.\\
@@
 
-	\paragraph{Error covariance matrices} : as mentioned above, these matrices set can be estimated with good accuracy using ALS algorithms [CITATION NEEDED].
+	\paragraph{Error covariance matrices} : as mentioned above, these matrices set can be estimated with good accuracy using ALS algorithms \cite{ALS}.
 It is to be noted that errors on these matrices will not induce new errors in the estimated images, as they are only used to compute the Kalman gain.
@@
 	We can then get the mean error values, measured using a distance defined above, and standard deviations needed.
+
+	\subsection{Initialisation error}
+
+	When using the Kalman filter, having the most accurate initialisation is important.
+This part has two goals: measure how an initialisation error affects reconstruction; and verify that the filter works correctly by eliminating these errors after a few snapshots.
+\\
+	To measure this, every parameter is considered perfect except the initial $\x_0$ and $\P_0$ computed accordingly.
+This test was done using a $3\times 3$ image of a \mintinline{matlab}|linspace|\footnote{This was done before the methodology establishment}, to which a $90\deg$ rotation was applied each snapshot.
+The error on the initial image was introduced by changing the Signal to Noise Ratio (SNR) during reconstruction.
+\\
+
+	The following reconstruction can be observed :
+
+	\begin{figure}[H]
+		\centering
+		\includegraphics[width=.8\linewidth]{src/K_ev}
+		\caption{Kalman filtering at different snapshots}
+		\label{fig:K_ev}
+	\end{figure}
+
+	On this figure (\ref{fig:K_ev}), the top line represents the original images set with the first replaced with inaccurate reconstruction, and the bottom line is the Kalman reconstruction at the same snapshots, titles of bottom images are the PSNR of these measures.
+\\
+
+	This figure illustrates how the initialisation error spreads between snapshots and how the filtering rapidly compensate it. This evolution can be plotted :
+
+	\begin{figure}[H]
+		\centering
+		\includegraphics[width=.5\linewidth]{src/evPSNR}
+		\caption{PSNR evolution of 1 simulation with inaccurate initialisation}
+	\end{figure}
+
+	On this figure, it can be observed that the error is compensated after 4 iterations. \\
+
+	To determine the number of snapshots in which the initialisation error was not the most corrected, a set of measures was done using different SNR :
+
+	\begin{figure}[H]
+		\centering
+		\includesvg[width=.65\linewidth]{graphs/SNR.svg}
+		\caption{Number of iteration needed to reach the best PSNR}
+	\end{figure}
+
+	An important observation made during experimentation is that the number of snapshots needed was never observed over 5.
+It mostly stays around 4 and sometimes is 5, so this is the number of snapshots we will consider for stabilisation of initial error.
+\\
+	The measure of the maximum PSNR can also be measured, but was not in this work.
 
@@
 		\caption{Image reconstruction at unreliable distance}
+		\label{fig:high_err}
 	\end{figure}
@@
 To evaluate the reconstruction error, the $d_2$ distance is used, as it is additive: the total distance between two reconstructions is the sum of the distance of each direction.
+The maximum distance is thus not normalised and errors superior to 1 appear, these errors cannot be compared to directional errors above.
+The results are the following :
+
+	\begin{figure}[H]
+		\centering
+		\begin{subfigure}{.5\textwidth}
+			\centering
+			\includesvg[width=.9\linewidth]{graphs/dir4_tot.svg}
+			\caption{Distance to true image}
+		\end{subfigure}%
+		\begin{subfigure}{.5\textwidth}
+			\centering
+			\includesvg[width=.9\linewidth]{graphs/dir4_1.svg}
+			\caption{Distance of inaccurate direction to the true direction}
+		\end{subfigure}
+		\caption{Simulation results for directional inaccuracy on one direction}
+	\end{figure}
+
+	The first notable observation is that the standard deviation of reconstruction error under directional perturbation is important, even after averaging 100 simulations : the reconstruction is thus more sensitive to directional perturbation.
+Another observation is that past the 0.05 direction error, the average reconstruction error is basically the same, with very small variations of the confidence interval.
+This is in fact the same case as figure (\ref{fig:high_err}), where the error is only the mean distance between two random images as the reconstruction is not reliable any more, being mostly errors.
+\\
+
+	Another remarkable observation is that the error on the single direction seems to grow slightly faster than the total error.
+This can be verified when computing the average percentage of the single direction error in the total one :
+
+	\begin{figure}[H]
+		\centering
+		\includesvg[width=.65\linewidth]{graphs/dir4_perc.svg}
+		\caption{Part of the single direction reconstruction error in the total one}
+	\end{figure}
+
+	This figure also shows that the reconstruction error in the erroneous direction is more important than on the others as it was expected (out of 100 directions, the mean error should be 1\% in each direction).
+This indicates that prior knowledge on the error may help identifying outliers in order to ignore them if their value makes no sense in regard of the other directions.
 
@@
 
+	To better understand the reconstruction error sensitivity to direction inaccuracies, the same simulation was done with errors on all directions, using $d_1$ distance for normalisation and reference to position error.
+The results are the following :
+
+	\begin{figure}[H]
+		\centering
+		\begin{subfigure}{.5\textwidth}
+			\centering
+			\includesvg[width=.9\linewidth]{graphs/dirA_x_e.svg}
+			\caption{Distance to true image}
+		\end{subfigure}%
+		\begin{subfigure}{.5\textwidth}
+			\centering
+			\includesvg[width=.9\linewidth]{graphs/dirA_xx.svg}
+			\caption{Distance to unperturbed reconstruction}
+		\end{subfigure}
+		\caption{Simulation results for directional inaccuracy on all direction}
+	\end{figure}
+
+	It can be observed that the error tends faster towards the random error case seen in (\ref{fig:high_err}).
+This can be qualified as directional resolution was equal to $2/10=0.2$ (direction cosines) between two directions, and positional resolution was $1.472$m between two  antennas.
+With this taken into account, the reconstruction is still more sensitive to directional error than positional error.
+
 	\newpage
 	\part{Experimental correction}
-	\section{implementation}
-	\section{Measures}
+
+	The filter being sensitive to directional error, the goal of this part is to use prior knowledge to estimate this error.
+The idea is to use a known $\y$, measured with the inaccessible $\H$, and a MVDR reconstruction $\x_{MVDR}$ \cite{TER} from this measure and solve the problem :
+
+	% IDEA : tourner sans bouger x_MVDR
+	% => ça marche pas
+
+	$$
+		\y - \H^d*\x_{MVDR} = 0
+	$$
+	Where :
+
+	$$
+		\H_{j,q}^d = \exp(-i\frac{2\pi}{\lambda}\Delta\z_j\dotproduct(\I_q^* - \u_q))
+	$$
+
+	The resulting vector $\u$ should be the correction vector such as :
+	$$
+		\I = \I^*-\u
+	$$
+
+	This problem being not linear, its analytic resolution is complicated.
+To solve it, it was implemented under Matlab using \mintinline{matlab}|lsqnonlin| function.
+This function does not support complex values, so the problem had to be broken up into real and complex parts (solved simultaneously), see code details in appendix.\\
+
+	However, after numerous tests, solving this problem does not seem feasible, at least using this function and these parameters: all the tests presented a solution worse than simply keeping the inaccurate matrix.
+The average deterioration has been measured :
+
+	$$
+		\left\langle d_1(\I,\I^*) - d_1(\I,\I^*-\u)\right\rangle = -4.12\times 10^{-4}
+	$$
+
+	To alleviate this issue, introducing an \emph{a priori} on the initial solution has been tested, without notable amelioration: the deterioration was only slightly weaker. \\
+
+	Some other solutions that were tried without being able to solve the problem:
+	\begin{itemize}
+		\item lowering function tolerance
+		\item lowering/raising step size and tolerance
+		\item switch from trust region reflective algorithm to Levenberg-Maquardt
+		\item raising max iterations and max evaluations to allow the algorithm to reach solutions with these parameters
+	\end{itemize}
 
 	\newpage
+
 	\part{Conclusion}
 
+	The results of this study demonstrate multiple points regarding the usability of a Kalman filter in the radio astronomic context.
+The first conclusion is that errors in the initial picture, even in good conditions, can continue to spread errors up to the fifth snapshot.
+Then, antennas positions should be known with at least 0.5\% relative precision, as an inaccuracy superior to that will create non-negligible errors in the final reconstruction.
+Concerning directions, those must be known with at least 0.1\% relative precision for the same reason.
+\\
+
+	Trying to estimate the error on directions could be a good way to get better results, however, analytic solution or a different resolution algorithm should be found, as both algorithms (trust region reflective and Levenberg-Maquardt) used by \mintinline{matlab}|lsqnonlin| does not produce conclusive results.
+
 	\newpage
-	\part{Further insights}
+	\part{Recommended future works}
 
@@
 In particular, model based deep learning \cite{deep} can be an effective way of rebuilding images when an evolution model is known.
+
+	\newpage
+	\part{Appendices}
+	\section{Experimental correction code}
+
+\begin{minted}{matlab}
+% init
+x0 = zeros(1,2*D);
+lb = [-0.1*ones(1,2*D)];
+ub = [0.1*ones(1,2*D)];
+
+% problem definition
+Iv = reshape(I_err,[],1); % I_err is I with errors
+prb = @(x)prb_f(x,y,Iv,dz,J,D,lambda,X_mvdr); % creation of function f(x) = y-H(x)*x_MVDR with static inputs
+
+% lsqnonlin
+options = optimoptions(@lsqnonlin,'Display','Iter','FiniteDifferenceType','central'); % FiniteDifferenceType central computes on both sides of the gradient, did not help but a bit more precise
+x = lsqnonlin(prb,x0,lb,ub,options); % solve f(x)=0
+
+% result extraction
+Icorr = - reshape(x(1:2*D),D,2) + I_err; % I_corr is I corrected
+Hcorr = matF(J,D,z_err,lambda,Icorr); % matF computes H
+
+.
+.
+.
+
+% function definition at the end of the file
+function s = prb_f(x,y,Iv,dz,J,D,lambda,xMVDR)
+Hmodif = zeros(J^2,D);
+for q=1:D
+Hmodif(: ,q) = exp(-1i*2*pi/lambda*(dz(:,1)*(Iv(q) - x(q))+dz(:,2)*(Iv(D+q) - x(D+q))));
+end 
+ymodif = Hmodif*xMVDR;
+s = [real(y);imag(y)] - [real(ymodif);imag(ymodif)];
+end
+
+\end{minted}
 

### 2022-08-24 main_loop.m

@@ -110,16 +110,21 @@ for err = 1:npos
         
         Iv = reshape(I_err,[],1);
 
-        prb = @(x)prb_f(x,y,Iv,dz,J,D,lambda);
+        prb = @(x)prb_f(x,y,Iv,dz,J,D,lambda,X_mvdr);
 
 
         % 2. init
-        dbiasPrio = zeros(1,2*D)*err/100;
-        xparfait = I_err-I;
-        dbiasPrioInit = reshape(xparfait,1,2*D) + dbiasPrio;
-        x0 = [dbiasPrioInit reshape(X_mvdr,1,[])];
-        lb = [-0.1*ones(1,2*D) zeros(1,D)];
-        ub = [0.1*ones(1,2*D) ones(1,D)];
+%         dbiasPrio = zeros(1,2*D)*err/100;
+%         xparfait = I_err-I;
+%         dbiasPrioInit = reshape(xparfait,1,2*D) + dbiasPrio;
+%         x0 = [dbiasPrioInit reshape(X_mvdr,1,[])];
+%         lb = [-0.1*ones(1,2*D) zeros(1,D)];
+%         ub = [0.1*ones(1,2*D) ones(1,D)];
+
+        x0 = zeros(1,2*D);
+        lb = [-0.1*ones(1,2*D)];
+        ub = [0.1*ones(1,2*D)];
+
 
         % 3. lsqnonlin
         options = optimoptions(@lsqnonlin,'Display','Iter','FiniteDifferenceType','central');
@@ -128,7 +133,7 @@ for err = 1:npos
         x = lsqnonlin(prb,x0,lb,ub,options);
 
         % 4. nouveau X_0, calcul Hcorr
-        X_0 = x(2*D+1:3*D);
+%         X_0 = x(2*D+1:3*D);
         Icorr = - reshape(x(1:2*D),D,2) + I_err;
         Hcorr = matF(J,D,z_err,lambda,Icorr);
         
@@ -173,14 +178,14 @@ for err = 1:npos
 %     save(nom,"wK","wK_err","wX","wX_err","e_x","e_K","e_x_e","de_x","de_K","de_x_e","e_th","de_xx","e_xx"); % ajouter Xp Xp_e
 end % end err
 
[…]

### 2022-08-24 article/article.tex
complete ig

@@
-\documentclass[titlepage]{article}
+\documentclass[titlepage,11pt]{article}
 \usepackage[upright]{fourier}
@@
 \setlength\parindent{0pt}
+\fontsize{11pt}{14pt}\selectfont
 
@@
 	\LARGE
-	G. Robert-Dautun
+	G. Robert-Dautun\\
+	\vspace*{22px}
+	\normalsize
+	This work follows the TER work \cite{TER}. All notations and results should be included but I recommend reading the report prior to reading this work. (see ref) \\
 
@@
 	\end{titlepage}
+	\section*{Acknowledgements}
+	\vspace*{\fill}
+	Major thanks to I. Vin, who suggested the topic and supervised me during the TER and the internship.\\
+
+	Thanks to M. Marchesseau for collaborating on the TER. \\
+
+	Thanks to the SATIE lab for welcoming me for this internship.
+	\vspace*{\fill}
+	%
+
+	\newpage
+	\begin{abstract}
+		Radio astronomy use multiple sensors to observe distant cosmic objects.
+These objects are not fixed in time, they are constantly evolving, and models based on modern astrophysics allow the prediction of this evolution.
+A Kalman filter creates more likely output based on these observations and models.
+This study test whether this filter can be used in radio astronomy or not.
+Different scenarios were computed, introducing different inaccuracies that can happen in the real world, with a focus on ionospheric perturbations, and Kalman filter performances were assessed in these conditions.
+This study shows that errors on initialisation are partially corrected within a few snapshots, and that position and direction must be known with at least 0.5\% relative precision.
+A solution to estimate the introduced error to correct it before filtering was tested but did not produce notable results.
+A Kalman filter can thus be used for radio astronomy as long as antennas positions and directions and ionospheric perturbations are well known.
+This study does not explore analytic resolution for error estimation and does not use deep learning for image enhancement.
+	\end{abstract}
 
@@
 	I did my internship in the SIAME/MOSS division. \\
-	The SIAME/MOSS division is focused on getting a better understanding and mastering complex systems. The three main topics of the division are :
+	The SIAME/MOSS division is focused on getting a better understanding and mastering complex systems. The three main topics of the division are:
 	\paragraph{Data and image analysis:} this topic's main goal is to estimate physical or semantic quantities from observations data.
@@
 They develop tools and methods for multi-sensor instrumentation, embedded computing architectures and data flux synchronisation.
-Their work is divided in 4 points :
 
@@
 When using optical telescopes, the determining factor for resolution is the size of the primary mirror.
-When observing really distant objects, the required size to have a decent resolution becomes unrealistically big.
+When observing really distant objects, the required size to have a decent resolution becomes unrealistically large.
 To address this problem, radio interferometry replaces a single optical instrument with multiple antennas collecting electromagnetic waves.
@@
 
-	A Kalman filter is a linear quadratic estimator, using observations, state-transition model and knowledge about noise and errors to compute a more accurate result than observations alone.
+	A Kalman filter is a linear quadratic estimator, using observations, state-transition model and knowledge about noise and errors to compute a more likely result than observations alone.
 Radio astronomical measures being prone to noise and errors, usage of this filtering method seems \emph{a priori} a good way to improve accuracy of reconstruction.
@@
 	A sensitive part of the filter is the observations model, needing precise knowledge of antennas positions and directions.
-A of this goal work is to determine how inaccuracies in these measurements affect image reconstruction, and if it is realistically usable in radio astronomical imagery.
+A goal of this work is to determine how inaccuracies in these measurements affect image reconstruction, and if it is realistically usable in radio astronomical imagery.
 \\
@@
 	In order to simulate data, an image is introduced as the "reference image", this image will only be used in order to compute simulated observations.
-The selected picture must respect the following conditions :
+The selected picture must respect the following conditions:
 	\begin{itemize}
@@
 
-	When reviewing Kalman filter performances under real conditions, the reference image is not considered as knwon, and we need to reconstruct the image from the visibilities matrix.
+	When reviewing Kalman filter performances under real conditions, the reference image is not considered as known, and we need to reconstruct the image from the visibilities matrix.
 While different methods exist \cite{TER}, the ones used here are MVDR reconstructions followed by Maximum Entropy deconvolution.
@@
 The convergence at non-zero distance of the errorless reconstruction is also unexplained.
+\\
+
+	A relative precision of 0.5\% is then the minimum to obtain a result almost unaffected by the lack of knowledge in the antennas positions.
 
@@
 With this taken into account, the reconstruction is still more sensitive to directional error than positional error.
+\\
+
+	To obtain results without too much additional error, a 0.1\% relative precision is thus needed. As it becomes a simple translation of the positional errors graph.
 
@@
 	Trying to estimate the error on directions could be a good way to get better results, however, analytic solution or a different resolution algorithm should be found, as both algorithms (trust region reflective and Levenberg-Maquardt) used by \mintinline{matlab}|lsqnonlin| does not produce conclusive results.
+\\
+
+	The Kalman filter can thus be used in radio astronomy under the condition that positions, directions, and ionospheric perturbations are well-known.
 
@@
 
+	\paragraph{Model error :} This aspect has not been studied in this paper, and is an important lead to follow, as models are not perfect with limited input and knowledge of certain phenomena.
+
 	\newpage
@@
 Iv = reshape(I_err,[],1); % I_err is I with errors
-prb = @(x)prb_f(x,y,Iv,dz,J,D,lambda,X_mvdr); % creation of function f(x) = y-H(x)*x_MVDR with static inputs
+prb = @(x)prb_f(x,y,Iv,dz,J,D,lambda,X_mvdr); % creation of function f(x) = y-H(x)*x_MVDR 
+% with static inputs
 
 % lsqnonlin
-options = optimoptions(@lsqnonlin,'Display','Iter','FiniteDifferenceType','central'); % FiniteDifferenceType central computes on both sides of the gradient, did not help but a bit more precise
+options = optimoptions(@lsqnonlin,'Display','Iter','FiniteDifferenceType','central'); 
+% FiniteDifferenceType central computes on both sides of the gradient,
+% did not help but a bit more precise
 x = lsqnonlin(prb,x0,lb,ub,options); % solve f(x)=0
@@
 \end{minted}
+
+	The full code can be found \href{https://github.com/RUValaS/RDR}{here}.
 
